### 2019-08-13 paper/paper.tex
Add paper outline and update paper

\documentclass[conference]{IEEEtran}
\IEEEoverridecommandlockouts
% The preceding line is only needed to identify funding in the first footnote. If that is unneeded, please comment it out.
\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xcolor}
\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
\begin{document}

\title{Speech Emotion Detection Using 2D CNNs}

\author{\IEEEauthorblockN{Brian Pho, Thomas Truong, Svetlana Yanushkevich}
\IEEEauthorblockA{\textit{Department of Electrical and Computer Engineering}\\
\textit{University of Calgary}\\
Calgary, Alberta, Canada\\
\{brian.pho, thomas.truong, syanshk\}@ucalgary.ca}
}

\maketitle

\begin{abstract}
This document is a model and instructions for \LaTeX.
This and the IEEEtran.cls file define the components of your paper [title, text, heads, etc.]. *CRITICAL: Do Not Use Symbols, Special Characters, Footnotes, 
or Math in Paper Title or Abstract.
\end{abstract}

\begin{IEEEkeywords}
component, formatting, style, styling, insert
\end{IEEEkeywords}

\section{Introduction}
This document is a model and instructions for \LaTeX.
Please observe the conference page limits.

\section{Ease of Use}

\subsection{Maintaining the Integrity of the Specifications}

The IEEEtran class file is used to format your paper and style the text. All margins, 
column widths, line spaces, and text fonts are prescribed; please do not 
alter them. You may note peculiarities. For example, the head margin
measures proportionately more than is customary. This measurement 
and others are deliberate, using specifications that anticipate your paper 
as one part of the entire proceedings, and not as an independent document. 
Please do not revise any of the current designations.

\section{Prepare Your Paper Before Styling}
Before you begin to format your paper, first write and save the content as a 
separate text file. Complete all content and organizational editing before 
formatting. Please note sections \ref{AA}--\ref{SCM} below for more information on 
proofreading, spelling and grammar.

Keep your text and graphic files separate until after the text has been 
formatted and styled. Do not number text heads---{\LaTeX} will do that 
for you.

\subsection{Abbreviations and Acronyms}\label{AA}
Define abbreviations and acronyms the first time they are used in the text, 
even after they have been defined in the abstract. Abbreviations such as 
IEEE, SI, MKS, CGS, ac, dc, and rms do not have to be defined. Do not use 
abbreviations in the title or heads unless they are unavoidable.

\subsection{Units}
\begin{itemize}
\item Use either SI (MKS) or CGS as primary units. (SI units are encouraged.) English units may be used as secondary units (in parentheses). An exception would be the use of English units as identifiers in trade, such as ``3.5-inch disk drive''.
\item Avoid combining SI and CGS units, such as current in amperes and magnetic field in oersteds. This often leads to confusion because equations do not balance dimensionally. If you must use mixed units, clearly state the units for each quantity that you use in an equation.
\item Do not mix complete spellings and abbreviations of units: ``Wb/m\textsuperscript{2}'' or ``webers per square meter'', not ``webers/m\textsuperscript{2}''. Spell out units when they appear in text: ``. . . a few henries'', not ``. . . a few H''.
\item Use a zero before decimal points: ``0.25'', not ``.25''. Use ``cm\textsuperscript{3}'', not ``cc''.)
\end{itemize}

\subsection{Equations}
Number equations consecutively. To make your 
equations more compact, you may use the solidus (~/~), the exp function, or 
appropriate exponents. Italicize Roman symbols for quantities and variables, 
but not Greek symbols. Use a long dash rather than a hyphen for a minus 
sign. Punctuate equations with commas or periods when they are part of a 
sentence, as in:
\begin{equation}
a+b=\gamma\label{eq}
\end{equation}

Be sure that the 
symbols in your equation have been defined before or immediately following 
the equation. Use ``\eqref{eq}'', not ``Eq.~\eqref{eq}'' or ``equation \eqref{eq}'', except at 
the beginning of a sentence: ``Equation \eqref{eq} is . . .''

\subsection{\LaTeX-Specific Advice}

Please use ``soft'' (e.g., \verb|\eqref{Eq}|) cross references instead
of ``hard'' references (e.g., \verb|(1)|). That will make it possible
to combine sections, add equations, or change the order of figures or
citations without having to go through the file line by line.

Please don't use the \verb|{eqnarray}| equation environment. Use
\verb|{align}| or \verb|{IEEEeqnarray}| instead. The \verb|{eqnarray}|
environment leaves unsightly spaces around relation symbols.

Please note that the \verb|{subequations}| environment in {\LaTeX}
will increment the main equation counter even when there are no
equation numbers displayed. If you forget that, you might write an
article in which the equation numbers skip from (17) to (20), causing
the copy editors to wonder if you've discovered a new method of
counting.

{\BibTeX} does not work by magic. It doesn't get the bibliographic
data from thin air but from .bib files. If you use {\BibTeX} to produce a
bibliography you must send the .bib files.

{\LaTeX} can't read your mind. If you assign the same label to a
subsubsection and a table, you might find that Table I has been cross
referenced as Table IV-B3.

{\LaTeX} does not have precognitive abilities. If you put a
\verb|\label| command before the command that updates the counter it's
supposed to be using, the label will pick up the last counter to be
cross referenced instead. In particular, a \verb|\label| command
should not go before the caption of a figure or a table.

Do not use \verb|\nonumber| inside the \verb|{array}| environment. It
will not stop equation numbers inside \verb|{array}| (there won't be
any anyway) and it might stop a wanted equation number in the
surrounding equation.

\subsection{Some Common Mistakes}\label{SCM}
\begin{itemize}
\item The word ``data'' is plural, not singular.
\item The subscript for the permeability of vacuum $\mu_{0}$, and other common scientific constants, is zero with subscript formatting, not a lowercase letter ``o''.
\item In American English, commas, semicolons, periods, question and exclamation marks are located within quotation marks only when a complete thought or name is cited, such as a title or full quotation. When quotation marks are used, instead of a bold or italic typeface, to highlight a word or phrase, punctuation should appear outside of the quotation marks. A parenthetical phrase or statement at the end of a sentence is punctuated outside of the closing parenthesis (like this). (A parenthetical sentence is punctuated within the parentheses.)
\item A graph within a graph is an ``inset'', not an ``insert''. The word alternatively is preferred to the word ``alternately'' (unless you really mean something that alternates).
\item Do not use the word ``essentially'' to mean ``approximately'' or ``effectively''.
\item In your paper title, if the words ``that uses'' can accurately replace the word ``using'', capitalize the ``u''; if not, keep using lower-cased.
\item Be aware of the different meanings of the homophones ``affect'' and ``effect'', ``complement'' and ``compliment'', ``discreet'' and ``discrete'', ``principal'' and ``principle''.
\item Do not confuse ``imply'' and ``infer''.
\item The prefix ``non'' is not a word; it should be joined to the word it modifies, usually without a hyphen.
\item There is no period after the ``et'' in the Latin abbreviation ``et al.''.
\item The abbreviation ``i.e.'' means ``that is'', and the abbreviation ``e.g.'' means ``for example''.
\end{itemize}
An excellent style manual for science writers is \cite{b7}.

\subsection{Authors and Affiliations}
\textbf{The class file is designed for, but not limited to, six authors.} A 
minimum of one author is required for all conference articles. Author names 
should be listed starting from left to right and then moving down to the 
next line. This is the author sequence that will be used in future citations 
and by indexing services. Names should not be listed in columns nor group by 
affiliation. Please keep your affiliations as succinct as possible (for 
example, do not differentiate among departments of the same organization).

\subsection{Identify the Headings}
Headings, or heads, are organizational devices that guide the reader through 
your paper. There are two types: component heads and text heads.

Component heads identify the different components of your paper and are not 
topically subordinate to each other. Examples include Acknowledgments and 
References and, for these, the correct style to use is ``Heading 5''. Use 
``figure caption'' for your Figure captions, and ``table head'' for your 
table title. Run-in heads, such as ``Abstract'', will require you to apply a 
style (in this case, italic) in addition to the style provided by the drop 
down menu to differentiate the head from the text.

Text heads organize the topics on a relational, hierarchical basis. For 
example, the paper title is the primary text head because all subsequent 
material relates and elaborates on this one topic. If there are two or more 
sub-topics, the next level head (uppercase Roman numerals) should be used 
and, conversely, if there are not at least two sub-topics, then no subheads 
should be introduced.

\subsection{Figures and Tables}
\paragraph{Positioning Figures and Tables} Place figures and tables at the top and 
bottom of columns. Avoid placing them in the middle of columns. Large 
figures and tables may span across both columns. Figure captions should be 
below the figures; table heads should appear above the tables. Insert 
figures and tables after they are cited in the text. Use the abbreviation 
``Fig.~\ref{fig}'', even at the beginning of a sentence.

\begin{table}[htbp]
\caption{Table Type Styles}
\begin{center}
\begin{tabular}{|c|c|c|c|}
\hline
\textbf{Table}&\multicolumn{3}{|c|}{\textbf{Table Column Head}} \\
\cline{2-4} 
\textbf{Head} & \textbf{\textit{Table column subhead}}& \textbf{\textit{Subhead}}& \textbf{\textit{Subhead}} \\
\hline
copy& More table copy$^{\mathrm{a}}$& &  \\
\hline
\multicolumn{4}{l}{$^{\mathrm{a}}$Sample of a Table footnote.}
\end{tabular}
\label{tab1}
\end{center}
\end{table}

\begin{figure}[htbp]
%\centerline{\includegraphics{fig1.png}}
\caption{Example of a figure caption.}
\label{fig}
\end{figure}

Figure Labels: Use 8 point Times New Roman for Figure labels. Use words 
rather than symbols or abbreviations when writing Figure axis labels to 
avoid confusing the reader. As an example, write the quantity 
``Magnetization'', or ``Magnetization, M'', not just ``M''. If including 
units in the label, present them within parentheses. Do not label axes only 
with units. In the example, write ``Magnetization (A/m)'' or ``Magnetization 
\{A[m(1)]\}'', not just ``A/m''. Do not label axes with a ratio of 
quantities and units. For example, write ``Temperature (K)'', not 
``Temperature/K''.

\section*{Acknowledgment}

The preferred spelling of the word ``acknowledgment'' in America is without 
an ``e'' after the ``g''. Avoid the stilted expression ``one of us (R. B. 
G.) thanks $\ldots$''. Instead, try ``R. B. G. thanks$\ldots$''. Put sponsor 
acknowledgments in the unnumbered footnote on the first page.

\section*{References}

Please number citations consecutively within brackets \cite{b1}. The 
sentence punctuation follows the bracket \cite{b2}. Refer simply to the reference 
number, as in \cite{b3}---do not use ``Ref. \cite{b3}'' or ``reference \cite{b3}'' except at 
the beginning of a sentence: ``Reference \cite{b3} was the first $\ldots$''

Number footnotes separately in superscripts. Place the actual footnote at 
the bottom of the column in which it was cited. Do not put footnotes in the 
abstract or reference list. Use letters for table footnotes.

Unless there are six authors or more give all authors' names; do not use 
``et al.''. Papers that have not been published, even if they have been 
submitted for publication, should be cited as ``unpublished'' \cite{b4}. Papers 
that have been accepted for publication should be cited as ``in press'' \cite{b5}. 
Capitalize only the first word in a paper title, except for proper nouns and 
element symbols.

For papers published in translation journals, please give the English 
citation first, followed by the original foreign-language citation \cite{b6}.

\begin{thebibliography}{00}
\bibitem{b1} G. Eason, B. Noble, and I. N. Sneddon, ``On certain integrals of Lipschitz-Hankel type involving products of Bessel functions,'' Phil. Trans. Roy. Soc. London, vol. A247, pp. 529--551, April 1955.
\bibitem{b2} J. Clerk Maxwell, A Treatise on Electricity and Magnetism, 3rd ed., vol. 2. Oxford: Clarendon, 1892, pp.68--73.
\bibitem{b3} I. S. Jacobs and C. P. Bean, ``Fine particles, thin films and exchange anisotropy,'' in Magnetism, vol. III, G. T. Rado and H. Suhl, Eds. New York: Academic, 1963, pp. 271--350.
\bibitem{b4} K. Elissa, ``Title of paper if known,'' unpublished.
\bibitem{b5} R. Nicole, ``Title of paper with only first word capitalized,'' J. Name Stand. Abbrev., in press.
\bibitem{b6} Y. Yorozu, M. Hirano, K. Oka, and Y. Tagawa, ``Electron spectroscopy studies on magneto-optical media and plastic substrate interface,'' IEEE Transl. J. Magn. Japan, vol. 2, pp. 740--741, August 1987 [Digests 9th Annual Conf. Magnetics Japan, p. 301, 1982].
\bibitem{b7} M. Young, The Technical Writer's Handbook. Mill Valley, CA: University Science, 1989.
\end{thebibliography}
\vspace{12pt}
\color{red}
IEEE conference templates contain guidance text for composing and formatting conference papers. Please ensure that all template text is removed from your conference paper prior to submission to the conference. Failure to remove the template text from your paper may result in your paper not being published.

\end{document}

### 2019-08-13 src/hyperparam_tuning.py
Add hyperparam tuning scratch file

@@ -1,0 +1,23 @@
+from src.neural_network import nn_constants as nnc
+from src.main import main
+from src.neural_network import nn_model
+
+
+def hyperparam_search():
+    hyperparams = {
+        "lr" : [0.1, 0.01],
+        "optimizer": [adam, rmsprop]
+    }
+
+    for params in hyperparams.items():
+        nnc.OPTIMIZER = params["optimzer"]
+        nnc.LOSS = params["loss"]
+        nnc.MODEL_SAVE_PATH = r"fdsafs"
+        main()
+
+    print("Complete search for hyperparams")
+
+
+if __name__ == "__main__":
+    hyperparam_search()
+

### 2019-08-14 src/database_processor/db_common.py
Fix db_common and update gitignore

@@ -81,7 +81,7 @@ def k_hot_encode_label(label):
 
     # Only count the emotions with the highest amount of votes
     k_hot_label = k_hot_label / np.max(k_hot_label)
-    k_hot_label = np.floor(k_hot_label, dtype=int)
+    k_hot_label = np.floor(k_hot_label).astype(int)
 
     # If they're all zero, then this sample doesn't fit with the set of labels
     # that we're considering so drop it

### 2019-08-14 src/hyperparam_tuning.py
Delete hyperparam_tuning script

@@ -1,23 +1,0 @@
-from src.neural_network import nn_constants as nnc
-from src.main import main
-from src.neural_network import nn_model
-
-
-def hyperparam_search():
-    hyperparams = {
-        "lr" : [0.1, 0.01],
-        "optimizer": [adam, rmsprop]
-    }
-
-    for params in hyperparams.items():
-        nnc.OPTIMIZER = params["optimzer"]
-        nnc.LOSS = params["loss"]
-        nnc.MODEL_SAVE_PATH = r"fdsafs"
-        main()
-
-    print("Complete search for hyperparams")
-
-
-if __name__ == "__main__":
-    hyperparam_search()
-

### 2019-08-20 charts/charts.js
Update charts

@@ -20,6 +20,8 @@ Chart.helpers.merge(Chart.defaults.global.plugins.datalabels, {
     color: 'white'
 });
 
+Chart.defaults.global.defaultFontSize = 15;
+
 // Standard colors
 window.chartColors = {
 	red: 'rgb(255, 99, 132)',
@@ -28,7 +30,7 @@ window.chartColors = {
 	green: 'rgb(75, 192, 192)',
 	blue: 'rgb(54, 162, 235)',
 	purple: 'rgb(153, 102, 255)',
-	grey: 'rgb(231,233,237)'
+	grey: 'rgb(148, 150, 153)'
 };
 
 // Number of samples in each emotion class
@@ -55,7 +57,8 @@ var myDoughnutChart = new Chart(eCPC, {
     options: Object.assign(options, {
         title: {
             display: true,
-            text: 'Emotions'
+            text: 'Emotions',
+            fontSize: 15
         }
     })
 });
@@ -67,7 +70,7 @@ var myDoughnutChart = new Chart(lTPC, {
     type: 'doughnut',
     data: {
         datasets: [{
-            data: [19884, 727, 1064, 45, 1],
+            data: [19884, 727, 1064],
             backgroundColor: [
                 window.chartColors.red, 
                 window.chartColors.orange, 
@@ -77,7 +80,7 @@ var myDoughnutChart = new Chart(lTPC, {
         }],
     
         // These labels appear in the legend and in the tooltips when hovering different arcs
-        labels: ["1", "2", "3", "4", "5"]
+        labels: ["1", "2", "3"]
     },
     options: Object.assign(options, {
         title: {

### 2019-08-20 src/database_processor/processed_db_stats.py
Add remove_label_type function

@@ -91,8 +91,27 @@ def count_samples_per_db(samples):
     print("Total:", sum(db_counts.values()))
 
 
+def remove_label_type(samples, label_types):
+    """
+    Removes samples from the processed database that have a certain label type.
+    Label types are the number of labels a given sample has such as "2".
+
+    :param samples: A list of sample filenames
+    :param label_types: A list of label types to filter/remove
+    """
+    for sample in samples:
+        label = dg.read_label(sample)
+        num_labels = np.sum(label)
+
+        if num_labels in label_types:
+            sample_path = os.path.join(dbc.PROCESS_DB_PATH, sample)
+            os.remove(sample_path)
+            print("Removed sample:", sample)
+
+
 def main():
     sample_fns = os.listdir(dbc.PROCESS_DB_PATH)
+    remove_label_type(sample_fns, [4, 5])
     calc_emotion_class(sample_fns)
     calc_label_type(sample_fns)
     count_samples_per_db(sample_fns)

### 2019-08-20 charts/charts.js
Update values for charts

@@ -1,11 +1,6 @@
 // Standard/global options
 var options = {
     responsive: false,
-    // animation: {
-    //     onComplete: function(animation){
-    //         document.querySelector('.savegraph').setAttribute('href', this.toBase64Image());
-    //     }
-    // }
 }
 
 // Standard size
@@ -20,6 +15,7 @@ Chart.helpers.merge(Chart.defaults.global.plugins.datalabels, {
     color: 'white'
 });
 
+// Set font size
 Chart.defaults.global.defaultFontSize = 15;
 
 // Standard colors
@@ -40,7 +36,7 @@ var myDoughnutChart = new Chart(eCPC, {
     type: 'doughnut',
     data: {
         datasets: [{
-            data: [5994, 6209, 1960, 1884, 4745, 3040, 883],
+            data: [5954, 6177, 1934, 1851, 4732, 3006, 876],
             backgroundColor: [
                 window.chartColors.red, 
                 window.chartColors.orange, 
@@ -57,8 +53,7 @@ var myDoughnutChart = new Chart(eCPC, {
     options: Object.assign(options, {
         title: {
             display: true,
-            text: 'Emotions',
-            fontSize: 15
+            text: 'Emotions'
         }
     })
 });
@@ -80,7 +75,7 @@ var myDoughnutChart = new Chart(lTPC, {
         }],
     
         // These labels appear in the legend and in the tooltips when hovering different arcs
-        labels: ["1", "2", "3"]
+        labels: ["One", "Two", "Three"]
     },
     options: Object.assign(options, {
         title: {
@@ -98,7 +93,7 @@ var myDoughnutChart = new Chart(dOPC, {
     type: 'doughnut',
     data: {
         datasets: [{
-            data: [7442, 10039, 1440, 2800],
+            data: [7414, 10021, 1440, 2800],
             backgroundColor: [
                 window.chartColors.red, 
                 window.chartColors.orange, 

### 2019-08-21 src/neural_network/nn_model.py
Update NN model

@@ -32,21 +32,26 @@ def build_model():
     # model.add(layers.Dropout(0.5))
     # regularizers.l1_l2(l1=0.01, l2=0.01)
 
-    model.add(layers.Conv2D(300, (3, 3), activation='relu',
+    model.add(layers.Conv2D(200, (3, 3), activation='relu',
                             input_shape=nnc.INPUT_SHAPE))
     model.add(layers.MaxPooling2D((2, 2)))
-    model.add(layers.Conv2D(300, (3, 3), activation='relu'))
-    model.add(layers.MaxPooling2D((2, 2)))
-    model.add(layers.Conv2D(250, (3, 3), activation='relu'))
-    model.add(layers.MaxPooling2D((2, 2)))
+    model.add(layers.BatchNormalization())
     model.add(layers.Conv2D(200, (3, 3), activation='relu'))
     model.add(layers.MaxPooling2D((2, 2)))
+    model.add(layers.BatchNormalization())
     model.add(layers.Conv2D(150, (3, 3), activation='relu'))
     model.add(layers.MaxPooling2D((2, 2)))
+    model.add(layers.BatchNormalization())
+    model.add(layers.Conv2D(100, (3, 3), activation='relu'))
+    model.add(layers.MaxPooling2D((2, 2)))
+    model.add(layers.BatchNormalization())
     model.add(layers.Flatten())
     model.add(layers.Dense(512, activation='relu',))
+    model.add(layers.Dropout(0.5))
+    model.add(layers.Dense(256, activation='relu',))
+    model.add(layers.Dropout(0.5))
     model.add(layers.Dense(512, activation='relu',))
-    model.add(layers.Dense(512, activation='relu',))
+    model.add(layers.Dropout(0.5))
     model.add(layers.Dense(7, activation='sigmoid'))
     model.compile(optimizer=nnc.OPTIMIZER, loss=nnc.LOSS, metrics=nnc.METRICS)
 

### 2019-08-21 charts/charts.js
Update paper and charts

@@ -16,7 +16,7 @@ Chart.helpers.merge(Chart.defaults.global.plugins.datalabels, {
 });
 
 // Set font size
-Chart.defaults.global.defaultFontSize = 15;
+Chart.defaults.global.defaultFontSize = 20;
 
 // Standard colors
 window.chartColors = {

### 2019-08-21 research/paper/paper.tex
Organize research files into their own folder

\documentclass[conference]{IEEEtran}
\IEEEoverridecommandlockouts

\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xcolor}
\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
\begin{document}

\title{Multi-Label Emotion Classification in Speech Using 2D CNNs}

\author{\IEEEauthorblockN{Brian Pho, Thomas Truong, Svetlana Yanushkevich}
\IEEEauthorblockA{\textit{Biometric Technologies Laboratory, Department of Electrical and Computer Engineering}\\
\textit{University of Calgary}, Canada\\
\{brian.pho, thomas.truong, syanshk\}@ucalgary.ca}
}

\maketitle

\begin{abstract}
Current emotion classification systems using machine learning often ignore the multi-label data that comes with databases. In this paper we question whether machine learning can be used to detect multiple emotions in speech. To answer this question, we created an aggregate database from four voice emotion-labeled databases and train a 2D convolutional neural network on it to determine if the model could recognize multiple emotions in a speech sample. The model was able to classify the samples with an accuracy of 50\%. While this result is better than the baseline, we suggest future improvements.
\end{abstract}

\begin{IEEEkeywords}
speech, emotion, classification, neural networks
\end{IEEEkeywords}

\section{Introduction}

\subsection{Literature Review}

The field of affective computing studies the development of systems that can recognize, interpret, process, and simulate human affects. One subfield of affective computing that we are interested in is speech emotion classification. Classifying speech based on its emotional content is applicable to the development of more user friendly kiosks, more interactive digital assistants such as Siri or Google Assistant, and allows for a better understanding of what parts of speech contribute to its emotional content. One method of classification that has demonstrated great success is deep learning. Specifically, the use of convolution neural networks have proven to be successful at classifying emotions in speech as shown by \cite{}. Current research has successfully applied deep learning to automate this process and achieves an accuracy of 90\% \cite.

However, these models only consider a single emotion per speech sample and are often only trained on one to two databases. This paper extends upon this work by considering the problem of multi-label emotion classification of speech using four databases. The IEMOCAP and CREMA-D databases both include multiple labels for each speech sample but the data is usually discarded by only considering the emotion with the majority votes. We argue that discarding the other emotions results in a less realistic model of emotion classification due to not matching human performance and because it sidesteps the problem of ambiguous samples.

\subsection{Summary}

We approach the problem of multi-label emotion classification by using 2D convolution neural networks to classify speech samples into multiple emotions. We collected four databases, two with mutli-labeled and single-labeled samples, and two with only single-labeled samples. We then combined these four databases into an aggregate database by processing all of the speech samples into log-Mel spectrograms. These log-Mel spectrograms were then fed into a seven layer neural network consisting of four convolutional layers and three dense layers. After training the neural network, the model achieved a 50\% test accuracy.

\section{Results}

\subsection{Method Summary}

To determine if a machine learning model could perform multi-label emotion classification on speech, we first need to get speech samples with their labeled emotion. We considered eight databases and chose four based on accessibility and the number of common emotions. Given the variability of emotions in each database and the constraint adding more databases increases the difficulty of controlling the consistency of the samples, we chose the following databases
\begin{itemize}
	\item IEMOCAP
	\item TESS
	\item RAVDESS
	\item CREMA-D
\end{itemize}
We specifically picked databases that were spoken in English to keep the language consistent.

Each sample from a database would flow through the same preprocessing steps to maintain consistency. The preprocessing steps are
\begin{enumerate}
	\item The sample starts as a raw waveform in the form of time series points specifying the amplitude at a certain time.
	\item we then pad or crop the sample to a desired length. In our case, we set a maximum length of 4.5 seconds. Samples shorter than this were zero-padded on the right tail. Samples longer than this were cropped to this length and the extra information was discarded.
	\item If the sample came from a database that we considered noisy, then a noise reduction function was run on the sample to reduce the noise. We consider the IEMOCAP and CREMA-D databases to be noisy.
	\item The sample is then converted into a log-Mel spectrogram using a short-time Fourier transform (STFT) and Mel scale equation. The phase information was discard as it does not seem to hold relevant information. \cite{}
	\item The final step is to normalize the spectrograms to have values between negative one and one. This was done by using a minmax scaling function.
\end{enumerate}

After all of the databases were processed this way, the final aggregate database was fed into a neural network for training. We constructed a seven layer neural network of four convolutional layers and three dense layers. The network architecture is shown in Figure \ref{label}. The training details follow.
\begin{itemize}
	\item We applied dropout to the dense layers and batch normalization to the convolutional layers to deal with overfitting.
	\item We shuffled the aggregate database to make each training batch more homogenous and thus prevent large gradient updates.
	\item We split the aggregate database into 80\% training, 10\% validation, and 10\% testing.
	\item We updated the model's hyperparameters based on the validation loss and accuracy to improve the model's accuracy and generalizability.
\end{itemize}
After training, the model was run on the testing set to get the final accuracy of the model. The final accuracy achieved was 50\%.

\subsection{Rationale}

\section{Discussion}

\subsection{Results Summary}

We created a neural network model that achieved a classification accuracy of 50\% on the problem of multi-label emotion classification. This model can successfully classify multiple emotions in a sample and while this is a good result, we can do better.

\subsection{Limitations}

\subsection{Future Work}

\section*{Acknowledgment}
Acknowledge PURE award. Test citation: \cite{N.Fahlgren}

\bibliographystyle{IEEEtran}
\bibliography{IEEEabrv, references}

\end{document}

### 2019-08-21 src/main.py
Set Numpy seed and update confusion matrix display

@@ -15,6 +15,8 @@ def main():
     """
     The main entry point of this program.
     """
+    # Used to keep the shuffling of samples consistent between runs
+    np.random.seed(0)
     # Get the samples (log-mel spectrograms in the form of .npy files)
     sample_fns = dg.get_sample_filenames()
     np.random.shuffle(sample_fns)  # Shuffle the dataset

### 2019-08-21 src/neural_network/nn_model.py

@@ -233,31 +233,41 @@ def visualize_confusion_matrix(model, test_gen):
         true_output, pred_output)
 
     # Display the confusion matrix for each emotion
-    labels = ["0", "1"]
+    labels = ["Absent", "Present"]
 
-    for emotion in range(0, emc.NUM_EMOTIONS):
-        em_cm = confusion_matrix[emotion]
+    for em_index in range(0, emc.NUM_EMOTIONS):
+        # Normalize the confusion matrix
+        em_cm = confusion_matrix[em_index]
+        em_cm = em_cm.astype('float') / em_cm.sum(axis=1)[:, np.newaxis]
 
         # Plot the confusion matrix
-        plt.clf()
-        plt.imshow(em_cm, interpolation='nearest')
-
-        # Create colorbar
-        plt.colorbar()
+        fig, ax = plt.subplots()
+        im = ax.imshow(em_cm, interpolation='nearest',
+                       cmap=plt.get_cmap("Blues"))
+        ax.figure.colorbar(im, ax=ax)
 
         # Set the x and y axis labels
-        plt.ylabel("True")
-        plt.xlabel("Predicted")
+        em_label = emc.INVERT_EMOTION_MAP[em_index]
+        ax.set(xticks=np.arange(em_cm.shape[1]),
+               yticks=np.arange(em_cm.shape[0]),
+               # ... and label them with the respective list entries
+               xticklabels=labels, yticklabels=labels,
+               title='Normalized Confusion Matrix for {}'.format(em_label),
+               ylabel='True label',
+               xlabel='Predicted label')
 
-        # Set the x and y tick values and tick labels
-        tick_marks = np.arange(len(labels))
-        plt.xticks(tick_marks, labels)
-        plt.yticks(tick_marks, labels)
+        # Rotate the tick labels and set their alignment.
+        plt.setp(ax.get_xticklabels(), rotation=45, ha="right",
+                 rotation_mode="anchor")
 
-        # Display each value in the matrix
-        for row in range(em_cm.shape[0]):
-            for col in range(em_cm.shape[1]):
-                plt.text(col, row, em_cm[row][col])
+        # Loop over data dimensions and create text annotations.
+        fmt = '.2f'
+        thresh = em_cm.max() / 2.
+        for i in range(em_cm.shape[0]):
+            for j in range(em_cm.shape[1]):
+                ax.text(j, i, format(em_cm[i, j], fmt),
+                        ha="center", va="center",
[…]

### 2019-08-21 src/audio_processor/spectrogram.py
Implement visualize_input_batch function

@@ -84,7 +84,7 @@ def display_melspecgram(melspecgram):
 
     :param melspecgram: The spectrogram data in the form of a 2D np.array.
     """
-    plt.pcolormesh(melspecgram)
+    plt.pcolormesh(melspecgram, cmap="magma")
     plt.title("Log-Mel Magnitude Spectrogram")
     plt.ylabel("Log-Mel Bins (mel)")
     plt.xlabel("Time (sample)")

### 2019-08-21 src/main.py

@@ -42,6 +42,9 @@ def main():
     valid_gen = dg.BatchGenerator(valid_samples)
     train_gen = dg.BatchGenerator(train_samples)
 
+    # # Visualize input batch to ensure the network is getting valid data
+    # nnm.visualize_input_batch(train_gen)
+
     # Create and train the model
     model = nnm.build_model()
     # model.load_weights(nnc.MODEL_SAVE_PATH)

### 2019-08-21 src/neural_network/nn_model.py

@@ -271,3 +271,21 @@ def visualize_confusion_matrix(model, test_gen):
 
         fig.tight_layout()
         plt.show()
+
+
+def visualize_input_batch(input_gen):
+    """
+    Visualizes a sample input batch to the neural network.
+
+    :param input_gen: A BatchGenerator object, can be train, valid, or test
+    """
+    # Get a batch from the generator
+    batch_inputs, batch_targets = input_gen[0]
+
+    plt_num = 1
+    for specgram in batch_inputs:
+        plt.subplot(4, 8, plt_num)
+        plt.pcolormesh(np.squeeze(specgram, axis=2), cmap="magma")
+        plt_num += 1
+
+    plt.show()

### 2019-08-21 research/paper/paper.tex
Update paper
Update paper with figures

@@
 \usepackage{graphicx}
+\graphicspath{{figures/}}
 \usepackage{textcomp}
@@
 
-\title{Multi-Label Emotion Classification in Speech Using 2D CNNs}
+\title{Multi-Label Speech Emotion Recognition Using 2D CNNs}
 
@@
 Current emotion classification systems using machine learning often ignore the multi-label data that comes with databases.
-In this paper we question whether machine learning can be used to detect multiple emotions in speech.
-To answer this question, we created an aggregate database from four voice emotion-labeled databases and train a 2D convolutional neural network on it to determine if the model could recognize multiple emotions in a speech sample.
+In this paper, we address the problem of whether machine learning can be used to detect multiple emotions in speech and how accurate they can identify the emotions present.
+We created an combined database from four voice emotion-labeled databases and trained a 2D convolutional neural network on it to determine if the model could recognize multiple emotions in a speech sample.
 The model was able to classify the samples with an accuracy of 50\%.
-While this result is better than the baseline, we suggest future improvements.
+This result shows that it is possible to apply machine learning to the problem of multi-label speech emotion recognition and to achieve a reasonable accuracy.
 \end{abstract}
@@
 We collected four databases, two with mutli-labeled and single-labeled samples, and two with only single-labeled samples.
-We then combined these four databases into an aggregate database by processing all of the speech samples into log-Mel spectrograms.
+We then combined these four databases into an combined database by processing all of the speech samples into log-Mel spectrograms.
 These log-Mel spectrograms were then fed into a seven layer neural network consisting of four convolutional layers and three dense layers.
@@
 
-\section{Results}
+\section{Method}
 
@@
 \begin{itemize}
-	\item IEMOCAP
-	\item TESS
-	\item RAVDESS
-	\item CREMA-D
+	\item IEMOCAP \cite{busso_2008}
+	\item TESS \cite{dupuis_2011}
+	\item RAVDESS \cite{livingstone_2018}
+	\item CREMA-D \cite{cao_2014}
 \end{itemize}
 We specifically picked databases that were spoken in English to keep the language consistent.
+We combined these four databases for training, validating, and testing the neural network.
+The combined database is detailed in figures \ref{combined_db_db_split}, \ref{combined_db_db_split}, \ref{combined_db_db_split}
+
+\begin{figure}[h!]
+	\centering
+	\hspace{6mm}
+	\includegraphics[width=\linewidth]{combined_db_db_split.png} 
+	\caption{The proportion of each database in the combined database.}
+	\label{fig:combined_db_db_split}
+\end{figure}
+
+\begin{figure}[h!]
+	\centering
+	\hspace{6mm}
+	\includegraphics[width=\linewidth]{combined_db_emotions.png} 
+	\caption{The proportion of each emotion in the combined database.}
+	\label{fig:combined_db_db_split}
+\end{figure}
+
+\begin{figure}[h!]
+	\centering
+	\hspace{6mm}
+	\includegraphics[width=\linewidth]{combined_db_label_types.png} 
+	\caption{The proportion of each label type in the combined database.}
+	\label{fig:combined_db_db_split}
+\end{figure}
 
@@
 
-After all of the databases were processed this way, the final aggregate database was fed into a neural network for training.
+\begin{figure}[h!]
+	\centering
+	\hspace{6mm}
+	\includegraphics[width=\linewidth]{high_level_dataflow_diagram.png}
+	\caption{A high level overview of the processing stages that a speech sample goes through.}
+	\label{fig:high_level_dataflow_diagram}
+\end{figure}
+
+After all of the databases were processed this way, the final combined database was fed into a neural network for training.
 We constructed a seven layer neural network of four convolutional layers and three dense layers.
@@
 	\item We applied dropout to the dense layers and batch normalization to the convolutional layers to deal with overfitting.
-	\item We shuffled the aggregate database to make each training batch more homogenous and thus prevent large gradient updates.
-	\item We split the aggregate database into 80\% training, 10\% validation, and 10\% testing.
+	\item We shuffled the combined database to make each training batch more homogenous and thus prevent large gradient updates.
+	\item We split the combined database into 80\% training, 10\% validation, and 10\% testing.
 	\item We updated the model's hyperparameters based on the validation loss and accuracy to improve the model's accuracy and generalizability.
 \end{itemize}
-After training, the model was run on the testing set to get the final accuracy of the model. The final accuracy achieved was 50\%.
 
-\subsection{Rationale}
+\begin{figure}[h!]
+	\centering
+	\hspace{6mm}
+	\includegraphics[width=\linewidth]{neural_network_architecture.png}
+	\caption{The detailed architecture of the neural network including its eight layers and convolution filter size.}
+	\label{fig:high_level_dataflow_diagram}
+\end{figure}
+
+\section{Results}
+
+After training, the model was evaluated on the testing set to get the final accuracy of the model. The final accuracy achieved was 50\%. Confusion matrices for each emotion are shown in Figure \ref{}.
 
@@
 
-\subsection{Limitations}
+\subsection{Limitations and Future Work}
 
-\subsection{Future Work}
+One limitation of this work was that we only accounted for seven emotions but we know that there are more emotions than this. \cite{} Another limitations was that 
+Here we present future improvements of our work 
+\begin{itemize}
+	\item Improving the recognition accuracy by using more sophisticated neural network architectures such as LSTMs or using more databases.
+	\item Incorporating the phase data from the STFT and testing to see if that improves the accuracy.
+	\item Testing a binary relevant approach to this multi-label problem and comparing it to this model.
+\end{itemize}
+
+\subsection{Conclusion}
+
+Overall, this paper presented a convolutional neural network model that achieves a 50\% accuracy on multi-class, multi-label speech emotion recognition using four combined databases.
+We obtained this result by transforming raw speech samples into log-Mel spectrograms using STFT and the Mel scale.
+These log-Mel spectrograms are then fed into a eight layer neural network for classification for training.
+When we have completed training the neural network, the model was evaluated and achieved as 50\% accuracy.
+While this result is a promising start, we suggest improvements that future work can build upon to improve the accuracy.
 
 \section*{Acknowledgment}
-Acknowledge PURE award. Test citation: \cite{N.Fahlgren}
+Acknowledge PURE award.
 

### 2019-08-22 research/paper/paper.tex
Fix figure reference

@@
 
+The paper is outlined as follows:
+\begin{itemize}
+	\item Section II details the methodology we use such as the how we combined four databases and the architecture of the neural network.
+	\item Section III details the results from training the neural network.
+	\item Section IV discusses the results in the context of the field and how future work can improve upon the work.
+\end{itemize}
+
 \section{Method}
@@
 \end{itemize}
+
 We specifically picked databases that were spoken in English to keep the language consistent.
 We combined these four databases for training, validating, and testing the neural network.
-The combined database is detailed in figures \ref{combined_db_db_split}, \ref{combined_db_db_split}, \ref{combined_db_db_split}
+The combined database is detailed in figures \ref{fig:combined_db_db_split}, \ref{fig:combined_db_emotions}, \ref{fig:combined_db_label_types}.
 
@@
 	\caption{The proportion of each emotion in the combined database.}
-	\label{fig:combined_db_db_split}
+	\label{fig:combined_db_emotions}
 \end{figure}
@@
 	\caption{The proportion of each label type in the combined database.}
-	\label{fig:combined_db_db_split}
+	\label{fig:combined_db_label_types}
 \end{figure}

### 2019-08-22 research/charts/charts.js
Update charts and images

@@ -16,7 +16,7 @@ Chart.helpers.merge(Chart.defaults.global.plugins.datalabels, {
 });
 
 // Set font size
-Chart.defaults.global.defaultFontSize = 20;
+Chart.defaults.global.defaultFontSize = 40;
 
 // Standard colors
 window.chartColors = {

### 2019-08-29 research/paper/paper.tex
Update paper and literature review
Add IEEEabrv
Update paper to near completion

--- 2019-08-21 research/paper/paper.tex
@@
 \usepackage{graphicx}
+\usepackage{dblfloatfix}
+\usepackage{caption}
 \graphicspath{{figures/}}
+\usepackage[strings]{underscore}
 \usepackage{textcomp}
-\usepackage{xcolor}
-\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
-    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
+\usepackage{multirow}
+\usepackage[table,xcdraw]{xcolor}
+\usepackage{pifont}
+\newcommand{\cmark}{\ding{51}}
+\newcommand{\xmark}{\ding{55}}
+\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08emT\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
+
 \begin{document}
 
-\title{Multi-Label Speech Emotion Recognition Using 2D CNNs}
+\title{Multi-Label Speech Emotion Recognition Using 2D Convolutional Neural Networks}
 
@@
 \IEEEauthorblockA{\textit{Biometric Technologies Laboratory, Department of Electrical and Computer Engineering}\\
-\textit{University of Calgary}, Canada\\
-\{brian.pho, thomas.truong, syanshk\}@ucalgary.ca}
-}
+University of Calgary, Canada\\\{brian.pho, thomas.truong, syanshk\}@ucalgary.ca}}
 
@@
 \begin{abstract}
-Current emotion classification systems using machine learning often ignore the multi-label data that comes with databases.
-In this paper, we address the problem of whether machine learning can be used to detect multiple emotions in speech and how accurate they can identify the emotions present.
-We created an combined database from four voice emotion-labeled databases and trained a 2D convolutional neural network on it to determine if the model could recognize multiple emotions in a speech sample.
-The model was able to classify the samples with an accuracy of 50\%.
+Current speech emotion recognition systems often overlook the multi-label data that comes with databases.
+In this paper, we address the problem of whether machine learning can be used to detect multiple emotions in speech.
+We created a combined database consisting of four voice emotion-labeled databases, and we used it to train a 2D convolutional neural network to determine if the model could recognize multiple emotions in a speech sample.
+The model was able to classify the samples with an accuracy of 52.57\%.
 This result shows that it is possible to apply machine learning to the problem of multi-label speech emotion recognition and to achieve a reasonable accuracy.
@@
 \begin{IEEEkeywords}
-speech, emotion, classification, neural networks
+multi-label, speech, emotion, recognition, neural networks
 \end{IEEEkeywords}
@@
 The field of affective computing studies the development of systems that can recognize, interpret, process, and simulate human affects.
-One subfield of affective computing that we are interested in is speech emotion classification.
-Classifying speech based on its emotional content is applicable to the development of more user friendly kiosks, more interactive digital assistants such as Siri or Google Assistant, and allows for a better understanding of what parts of speech contribute to its emotional content.
-One method of classification that has demonstrated great success is deep learning.
-Specifically, the use of convolution neural networks have proven to be successful at classifying emotions in speech as shown by \cite{}.
-Current research has successfully applied deep learning to automate this process and achieves an accuracy of 90\% \cite.
+A subfield of affective computing that we are interested in is speech emotion recognition (SER).
+SER is the problem of recognizing which emotions are present in speech and SER has been approached in two ways:
+\begin{itemize}
+	\item Feature engineering
+	\item Machine learning (ML)
+\end{itemize} 
+Feature engineering is a method of manually extracting desired parameters from the data for use in classification and recognition.
+For audio, examples of parameters include pitch, energy, and Mel-Frequency Cepstral Coefficients (MFCC).\cite{Rybka2013} Nassif et al.
+\cite{Nassif2019} found that most researchers used the MFCC parameter for deep learning models but also recommended other parameters such as Linear Predictive Coding.
 
-However, these models only consider a single emotion per speech sample and are often only trained on one to two databases.
-This paper extends upon this work by considering the problem of multi-label emotion classification of speech using four databases.
-The IEMOCAP and CREMA-D databases both include multiple labels for each speech sample but the data is usually discarded by only considering the emotion with the majority votes.
-We argue that discarding the other emotions results in a less realistic model of emotion classification due to not matching human performance and because it sidesteps the problem of ambiguous samples.
+Machine learning is a method of creating an artificial neural network that automatically extracts its own parameters for classification and recognition.
+Examples include support vector machines (SVM), convolutional neural networks (CNN), and recurrent neural networks (RNN).
+Both approaches are not exclusive and a hybrid approach of feature engineering and machine learning results in better classification accuracy.\cite{Nassif2019}
+
+A common example of combining both approaches is to convert an audio waveform into a spectrogram using the short-time Fourier transform (STFT) and to feed the spectrogram into a neural network.
+Papers such as \cite{Engel2019}, \cite{Chen2018}, \cite{Badshah2019}, and \cite{Zhao2019} have demonstrated that this approach is successful for processing audio and for recognizing emotions in speech.
+
+Current state-of-the-art speech emotion recognition has achieved an accuracy of 52.14\% for the case of speaker-independent, single-labeled emotions on the IEMOCAP database.\cite{Zhao2019} This result was achieved using a 2D CNN LSTM model where the audio waveform was converted into a log-Mel spectrogram and then fed into the model.
+
+However, this model and most other models only consider a single emotion per speech sample and are often only trained on one or two databases.
+Kim et al.
+\cite{Kim2018a} is an exception as they consider the multi-label case of emotion recognition but they did not use machine learning and they used visual data on top of audio data.
+This paper extends upon the state-of-the-art by considering the problem of multi-label speech emotion classification using four databases to train and test the neural network.
+
+The IEMOCAP \cite{busso_2008} and CREMA-D \cite{cao_2014} databases both include multiple labels for each speech sample but the data is discarded by considering the emotion with the majority of votes.
+We argue that discarding the non-majority votes results in a less realistic model of emotion classification due to not matching human performance and due to the loss of information by discarding votes.
 
@@
 
-We approach the problem of multi-label emotion classification by using 2D convolution neural networks to classify speech samples into multiple emotions.
-We collected four databases, two with mutli-labeled and single-labeled samples, and two with only single-labeled samples.
-We then combined these four databases into an combined database by processing all of the speech samples into log-Mel spectrograms.
+We approach the problem of multi-label emotion classification by using a 2D CNN to classify speech samples into multiple emotions.
+We collected four databases: two with multi-label and single-label samples, and two with only single-label samples.
+We then combined the four databases into an a single database by processing all of the speech samples into log-Mel spectrograms.
 These log-Mel spectrograms were then fed into a seven layer neural network consisting of four convolutional layers and three dense layers.
-After training the neural network, the model achieved a 50\% test accuracy.
+After training the neural network, the model achieves a 52.57\% test accuracy.
 
-\section{Method}
+This paper is outlined as follows:
+\begin{itemize}
+	\item Section II details the methodology we use such as the how we combined the four databases and describes the architecture of the neural network.
+	\item Section III describes the results from testing the neural network.
+	\item Section IV discusses the results in the context of the field and how the results can be improved upon.
+\end{itemize}
 
-\subsection{Method Summary}
+\bgroup
+\def\arraystretch{2}%  1 is the default, change whatever you need
+\begin{table*}
+	\resizebox{\textwidth}{!}{%
+		\begin{tabular}{|l|c|c|c|c|c|c|c|c|c|c|c|c|c|}
+			\hline
+			\multicolumn{1}{|c|}{} & \multicolumn{13}{c|}{\textbf{Emotions}} \\ \cline{2-14} 
+			\multicolumn{1}{|c|}{\multirow{-2}{*}{\textbf{Databases}}} & \textbf{Neutral} & \textbf{Anger} & \textbf{Disgust} & \textbf{Fear} & \textbf{Happy} & \textbf{Sad} & \textbf{Surprise} & \textbf{Calm} & \textbf{Excitement} & \textbf{Frustration} & \textbf{Amused} & \textbf{Sleepy} & \textbf{Bored} \\ \hline
+			\rowcolor[HTML]{F6F8FA} 
+			\textbf{IEMOCAP} & \cmark & \cmark & \cmark & \cmark & \cmark & \cmark & \cmark & \xmark & \cmark & \cmark & \xmark & \xmark & \xmark \\ \hline
+			\textbf{CREMA-D} & \cmark & \cmark & \cmark & \cmark & \cmark & \cmark & \xmark & \xmark & \xmark & \xmark & \xmark & \xmark & \xmark \\ \hline
+			\rowcolor[HTML]{F6F8FA} 
+			\textbf{TESS} & \cmark & \cmark & \cmark & \cmark & \cmark & \cmark & \cmark & \xmark & \xmark & \xmark & \xmark & \xmark & \xmark \\ \hline
+			\textbf{RAVDESS} & \cmark & \cmark & \cmark & \cmark & \cmark & \cmark & \cmark & \cmark & \xmark & \xmark & \xmark & \xmark & \xmark \\ \hline
+			\rowcolor[HTML]{F6F8FA} 
+			\textbf{MSP-IMPROV} & \cmark & \cmark & \cmark & \cmark & \cmark & \cmark & \cmark & \xmark & \xmark & \xmark & \xmark & \xmark & \xmark \\ \hline
+			\textbf{SAVEE} & \cmark & \cmark & \cmark & \cmark & \cmark & \cmark & \cmark & \xmark & \xmark & \xmark & \xmark & \xmark & \xmark \\ \hline
+			\rowcolor[HTML]{F6F8FA} 
+			\textbf{Emo-DB} & \cmark & \cmark & \cmark & \cmark & \cmark & \cmark & \xmark & \xmark & \xmark & \xmark & \xmark & \xmark & \cmark \\ \hline
+			\textbf{EmoV-DB} & \cmark & \cmark & \cmark & \xmark & \xmark & \xmark & \xmark & \xmark & \xmark & \xmark & \cmark & \cmark & \xmark \\ \hline
+		\end{tabular}%
+	}
+	\caption{Comparison of databases with their labeled emotions.}
+	\label{tab:db_emo_table}
+\end{table*}
+\egroup
 
-To determine if a machine learning model could perform multi-label emotion classification on speech, we first need to get speech samples with their labeled emotion.
-We considered eight databases and chose four based on accessibility and the number of common emotions.
-Given the variability of emotions in each database and the constraint adding more databases increases the difficulty of controlling the consistency of the samples, we chose the following databases
+\section{Methodology}
+
+\subsection{Preprocessing}
+
+To determine if a ML model can perform multi-label speech emotion recognition, we first need to obtain speech samples with their labeled emotion.
+We considered eight databases and chose four based on accessibility and the number of overlapping emotions.
+Table \ref{tab:db_emo_table} compares the set of emotions of each database which makes it easier to identify overlapping emotions.
+We chose four databases because it becomes more difficult to maintain consistency due to database variability.
+Databases can vary in
 \begin{itemize}
+	\item The set and number of labeled emotions
+	\item The number of labels per sample
+	\item The audio quality such as sampling rate and noise
+	\item The spoken language
+\end{itemize}
+Given this variability, we chose the following four databases
+\begin{enumerate}
 	\item IEMOCAP \cite{busso_2008}
@@
 	\item CREMA-D \cite{cao_2014}
-\end{itemize}
-We specifically picked databases that were spoken in English to keep the language consistent.
-We combined these four databases for training, validating, and testing the neural network.
-The combined database is detailed in figures \ref{combined_db_db_split}, \ref{combined_db_db_split}, \ref{combined_db_db_split}
+\end{enumerate}
 
-\begin{figure}[h!]
+To control for the set and number of labeled emotions, we consider the following seven emotions for recognition
+\begin{enumerate}
+	\item Neutral
+	\item Anger
+	\item Disgust
+	\item Fear
+	\item Happy
+	\item Sad
+	\item Surprise
+\end{enumerate}
+We chose these seven emotions due to them being considered basic emotions by \cite{Ekman1992} and due to these seven being the most common among all databases.
+
+To control for the number of labels per sample, we mixed the multi-labeled data from the IEMOCAP and CREMA-D databases with the single-labeled data from the TESS and RAVDESS databases.
+We chose to mix of single- and multi- labeled data to increase the number of samples that the model can learn from and because the single-labels can be considered as special cases of multi-labeled data.
+However, we removed samples that were labeled with four or five emotions because we consider samples with that many emotions to be ambiguous and because outliers hinder a neural networks ability to learn.
+Outliers can cause large gradient updates which prevents a model from converging.
+
+To control for the audio quality, we resampled all samples to 48 kHz, applied noise reduction to samples from the IEMOCAP and CREMA-D databases, and cropped all samples to 4.5 seconds.
+We used a kaiser filter for resampling and used spectral gating for noise reduction.
+We only applied noise reduction to the IEMOCAP and CREMA-D databases after listening to the audio.
+Lastly, we chose databases that are spoken in English to maintain language consistency.
+
+We combined these four databases into a combined database to train, validate, and test the neural network. The combined database is detailed in figure \ref{fig:combined_db}.
+
+\begin{figure*}[t]
 	\centering
 	\hspace{6mm}
-	\includegraphics[width=\linewidth]{combined_db_db_split.png} 
-	\caption{The proportion of each database in the combined database.}
-	\label{fig:combined_db_db_split}
-\end{figure}
+	\includegraphics[width=\textwidth]{combined_db.png} 
+	\caption{Proportions of each: database, emotion, and label types in the combined database.}
+	\label{fig:combined_db}
+\end{figure*}
 
-\begin{figure}[h!]
-	\centering
-	\hspace{6mm}
-	\includegraphics[width=\linewidth]{combined_db_emotions.png} 
-	\caption{The proportion of each emotion in the combined database.}
-	\label{fig:combined_db_db_split}
-\end{figure}
-
-\begin{figure}[h!]
-	\centering
-	\hspace{6mm}
-	\includegraphics[width=\linewidth]{combined_db_label_types.png} 
-	\caption{The proportion of each label type in the combined database.}
-	\label{fig:combined_db_db_split}
-\end{figure}
-
-Each sample from a database would flow through the same preprocessing steps to maintain consistency. The preprocessing steps are
+Each sample from a database would flow through the same preprocessing steps to maintain consistency across all samples.
+A high level overview of the preprocessing steps is shown in figure \ref{fig:high_level_dataflow_diagram} and the detailed steps are described below.
 \begin{enumerate}
-	\item The sample starts as a raw waveform in the form of time series points specifying the amplitude at a certain time.
-	\item we then pad or crop the sample to a desired length.
-In our case, we set a maximum length of 4.5 seconds.
-Samples shorter than this were zero-padded on the right tail.
-Samples longer than this were cropped to this length and the extra information was discarded.
-	\item If the sample came from a database that we considered noisy, then a noise reduction function was run on the sample to reduce the noise.
-We consider the IEMOCAP and CREMA-D databases to be noisy.
-	\item The sample is then converted into a log-Mel spectrogram using a short-time Fourier transform (STFT) and Mel scale equation.
+	\item A sample starts as a raw waveform in the form of time series points specifying the amplitude at a certain point in time.
+	\item The sample is then padded or cropped to the desired length of 4.5s. Shorter samples were zero-padded on the right tail. Longer samples were cropped and the extra information discarded.
+	\item If the sample came from a database that we considered noisy, then a noise reduction filter was applied to the sample. We consider the IEMOCAP and CREMA-D databases to be noisy.
+	\item The sample is then converted into a log-Mel spectrogram using the short-time Fourier transform (STFT) and Mel scale equation.
 The phase information was discard as it does not seem to hold relevant information.
-\cite{}
+\cite{Kozakowski2017}
 	\item The final step is to normalize the spectrograms to have values between negative one and one. This was done by using a minmax scaling function.
@@
 
-\begin{figure}[h!]
+For the STFT, we used a window size of 3072 with a 75\% overlap. For the Mel spectrogram, we set the minimum frequency to 20 Hz and maximum frequency to 12 kHz with 200 Mel bins.
+
+\begin{figure}[]
 	\centering
@@
 
-After all of the databases were processed this way, the final combined database was fed into a neural network for training.
-We constructed a seven layer neural network of four convolutional layers and three dense layers.
-The network architecture is shown in Figure \ref{label}.
-The training details follow.
+\subsection{Neural Network}
+
+\begin{figure*}[b!]
+	\centering
+	\hspace{6mm}
+	\includegraphics[width=\textwidth]{neural_network_architecture.png}
+	\caption{Architecture for the speech emotion recognition network including the input layer and seven processing layers.}
+	\label{fig:neural_network_architecture}
+\end{figure*}
+
+After all of the databases were processed this way, the final combined database was fed into a neural network for training, validation, and testing.
+We constructed a seven layer neural network consisting of four convolutional layers and three dense layers.
+The network architecture is shown in figure \ref{fig:neural_network_architecture}.
+The training process is described below.
 \begin{itemize}
-	\item We applied dropout to the dense layers and batch normalization to the convolutional layers to deal with overfitting.
-	\item We shuffled the combined database to make each training batch more homogenous and thus prevent large gradient updates.
+	\item We shuffled the combined database to make each input batch more uniform thus mitigating large gradient updates.
 	\item We split the combined database into 80\% training, 10\% validation, and 10\% testing.
-	\item We updated the model's hyperparameters based on the validation loss and accuracy to improve the model's accuracy and generalizability.
+	\item We applied dropout \cite{Srivastava2014} to the dense layers and batch normalization \cite{Ioffe2015} to the convolutional layers to deal with overfitting.
+	\item We updated the model's hyperparameters based on the validation loss and accuracy to improve the model's accuracy and ability to generalize.
+	\item We used class weights during training to address class imbalance.
 \end{itemize}
 
-\begin{figure}[h!]
-	\centering
-	\hspace{6mm}
-	\includegraphics[width=\linewidth]{neural_network_architecture.png}
-	\caption{The detailed architecture of the neural network including its eight layers and convolution filter size.}
-	\label{fig:high_level_dataflow_diagram}
-\end{figure}
+Table \ref{tab:model_hyperparams} describes the network hyperparameters and table \ref{tab:train_hyperparams} describes the training hyperparameters.
+
+\begin{table}[b]
+	\resizebox{\linewidth}{!}{%
+		\begin{tabular}{|l|r|}
+			\hline
+			\textbf{Hyperparameter} & \textbf{Value} \\ \hline
+			\rowcolor[HTML]{F6F8FA} 
+			Input Dimensions & 278w x 200h x 1d \\ \hline
+			Optimization Algorithm & Adam \\ \hline
+			\rowcolor[HTML]{F6F8FA} 
+			Loss Measure & Binary Cross-entropy \\ \hline
+			Accuracy Metric & Categorical Cross-entropy \\ \hline
+			\rowcolor[HTML]{F6F8FA} 
+			Activation Function & Sigmoid \\ \hline
+		\end{tabular}%
+	}
+	\caption{The model's hyperparameters.}
+	\label{tab:model_hyperparams}
+\end{table}
+
+\begin{table}[]
+	\resizebox{\linewidth}{!}{%
+		\begin{tabular}{|l|r|}
+			\hline
+			\textbf{Hyperparameter} & \textbf{Value} \\ \hline
+			\rowcolor[HTML]{F6F8FA} 
+			Epochs & 20 \\ \hline
+			Batch Size & 32 \\ \hline
+			\rowcolor[HTML]{F6F8FA} 
+			Training Set & 17,341 \\ \hline
+			Validation Set & 2,167 \\ \hline
+			\rowcolor[HTML]{F6F8FA} 
+			Testing Set & 2,167 \\ \hline
+		\end{tabular}%
+	}
+	\caption{The training hyperparameters.}
+	\label{tab:train_hyperparams}
+\end{table}
 
@@
 
-After training, the model was evaluated on the testing set to get the final accuracy of the model. The final accuracy achieved was 50\%. Confusion matrices for each emotion are shown in Figure \ref{}.
+After training the model, the model was evaluated on the testing set to get the final accuracy of the model.
+The final accuracy achieved was 52.57\%.
+Confusion matrices for each emotion are shown in Figure \ref{fig:confusion_matrix}.
+
+\begin{figure}[]
+	\centering
+	\hspace{6mm}
+	\includegraphics[width=\textwidth]{confusion_matrix.png}
+	\caption{Confusion matrices for each emotion.}
+	\label{fig:confusion_matrix}
+\end{figure}
 
@@
 
-\subsection{Results Summary}
-
-We created a neural network model that achieved a classification accuracy of 50\% on the problem of multi-label emotion classification.
+We created a neural network model that achieved a classification accuracy of 52.57\% on the problem of multi-label emotion classification.
 This model can successfully classify multiple emotions in a sample and while this is a good result, we can do better.
@@
 
-Overall, this paper presented a convolutional neural network model that achieves a 50\% accuracy on multi-class, multi-label speech emotion recognition using four combined databases.
+Overall, this paper presented a convolutional neural network model that achieves a 52.57\% accuracy on multi-class, multi-label speech emotion recognition using four combined databases.
 We obtained this result by transforming raw speech samples into log-Mel spectrograms using STFT and the Mel scale.
 These log-Mel spectrograms are then fed into a eight layer neural network for classification for training.
-When we have completed training the neural network, the model was evaluated and achieved as 50\% accuracy.
+When we have completed training the neural network, the model was evaluated and achieved as 52.57\% accuracy.
 While this result is a promising start, we suggest improvements that future work can build upon to improve the accuracy.
@@
 \bibliographystyle{IEEEtran}
-\bibliography{IEEEabrv, references}
+\bibliography{IEEEabrv,references}
 

### 2019-08-30 research/paper/paper.tex
Complete paper draft

@@
 \usepackage{dblfloatfix}
+\usepackage{hyperref}
 \usepackage{caption}
-\graphicspath{{figures/}}
-\usepackage[strings]{underscore}
+\usepackage{pifont}
 \usepackage{textcomp}
@@
 \usepackage[table,xcdraw]{xcolor}
-\usepackage{pifont}
+\usepackage[strings]{underscore}
+
+\graphicspath{{figures/}}
 \newcommand{\cmark}{\ding{51}}
@@
 \IEEEauthorblockA{\textit{Biometric Technologies Laboratory, Department of Electrical and Computer Engineering}\\
-University of Calgary, Canada\\\{brian.pho, thomas.truong, syanshk\}@ucalgary.ca}}
+University of Calgary, Canada\\
+\{brian.pho, thomas.truong, syanshk\}@ucalgary.ca}}
 
@@
 In this paper, we address the problem of whether machine learning can be used to detect multiple emotions in speech.
-We created a combined database consisting of four voice emotion-labeled databases, and we used it to train a 2D convolutional neural network to determine if the model could recognize multiple emotions in a speech sample.
+We created a combined database consisting of four speech emotion-labeled databases, and we used it to train a 2D convolutional neural network to determine if the model could recognize multiple emotions in a speech sample.
 The model was able to classify the samples with an accuracy of 52.57\%.
-This result shows that it is possible to apply machine learning to the problem of multi-label speech emotion recognition and to achieve a reasonable accuracy.
+This shows that it is possible to apply machine learning to the problem of multi-label speech emotion recognition and to achieve a reasonable accuracy.\footnote{\url{https://github.com/Brian-Pho/RVST598_Speech-Emotion-Recognition}}
 \end{abstract}
@@
 \end{itemize} 
-Feature engineering is a method of manually extracting desired parameters from the data for use in classification and recognition.
+Feature engineering is a method of manually extracting desired parameters from the data for use in  recognition.
 For audio, examples of parameters include pitch, energy, and Mel-Frequency Cepstral Coefficients (MFCC).\cite{Rybka2013} Nassif et al.
@@
 
-Machine learning is a method of creating an artificial neural network that automatically extracts its own parameters for classification and recognition.
+Machine learning is a method of creating an artificial neural network that automatically extracts its own parameters for recognition.
 Examples include support vector machines (SVM), convolutional neural networks (CNN), and recurrent neural networks (RNN).
@@
 
-A common example of combining both approaches is to convert an audio waveform into a spectrogram using the short-time Fourier transform (STFT) and to feed the spectrogram into a neural network.
+A common example of combining both approaches is to convert an audio waveform into a spectrogram using the Short-Time Fourier Transform (STFT) and to feed the spectrogram into a neural network.
 Papers such as \cite{Engel2019}, \cite{Chen2018}, \cite{Badshah2019}, and \cite{Zhao2019} have demonstrated that this approach is successful for processing audio and for recognizing emotions in speech.
 
-Current state-of-the-art speech emotion recognition has achieved an accuracy of 52.14\% for the case of speaker-independent, single-labeled emotions on the IEMOCAP database.\cite{Zhao2019} This result was achieved using a 2D CNN LSTM model where the audio waveform was converted into a log-Mel spectrogram and then fed into the model.
+Current state-of-the-art speech emotion recognition has achieved an accuracy of 52.14\% for the case of speaker-independent, single-labeled emotions on the IEMOCAP database.\cite{Zhao2019} This result was achieved using a 2D CNN long short-term memory (LSTM) model where the audio waveform was converted into a log-Mel spectrogram and then fed into the model.
 
@@
 We then combined the four databases into an a single database by processing all of the speech samples into log-Mel spectrograms.
-These log-Mel spectrograms were then fed into a seven layer neural network consisting of four convolutional layers and three dense layers.
+These log-Mel spectrograms were then fed into an eight-layer neural network consisting of four convolutional layers and four dense layers.
 After training the neural network, the model achieves a 52.57\% test accuracy.
@@
 \bgroup
-\def\arraystretch{2}%  1 is the default, change whatever you need
+\def\arraystretch{2}
 \begin{table*}
@@
 	\caption{Comparison of databases with their labeled emotions.}
-	\label{tab:db_emo_table}
+	\label{dbEmoTable}
 \end{table*}
@@
 We considered eight databases and chose four based on accessibility and the number of overlapping emotions.
-Table \ref{tab:db_emo_table} compares the set of emotions of each database which makes it easier to identify overlapping emotions.
+Table \ref{dbEmoTable} compares the set of emotions of each database which makes it easier to identify overlapping emotions.
 We chose four databases because it becomes more difficult to maintain consistency due to database variability.
@@
 \end{enumerate}
-We chose these seven emotions due to them being considered basic emotions by \cite{Ekman1992} and due to these seven being the most common among all databases.
+We chose these seven emotions due to them being considered basic emotions by Ekman \cite{Ekman1992} and due to these seven being the most common among all databases.
 
@@
 
-We combined these four databases into a combined database to train, validate, and test the neural network. The combined database is detailed in figure \ref{fig:combined_db}.
+We combined these four databases into a combined database to train, validate, and test the neural network. The combined database is detailed in figure \ref{combinedDb}.
 
@@
 	\includegraphics[width=\textwidth]{combined_db.png} 
-	\caption{Proportions of each: database, emotion, and label types in the combined database.}
-	\label{fig:combined_db}
+	\caption{Proportions of each database, emotion, and label types in the combined database.}
+	\label{combinedDb}
 \end{figure*}
@@
 Each sample from a database would flow through the same preprocessing steps to maintain consistency across all samples.
-A high level overview of the preprocessing steps is shown in figure \ref{fig:high_level_dataflow_diagram} and the detailed steps are described below.
+A high level overview of the preprocessing steps is shown in figure \ref{highLevelDataFlowDiagram} and the detailed steps are described below.
 \begin{enumerate}
@@
 \cite{Kozakowski2017}
-	\item The final step is to normalize the spectrograms to have values between negative one and one. This was done by using a minmax scaling function.
+	\item The final step is to normalize the spectrograms to have values between negative one and one. This was done by using a min-max scaling function.
 \end{enumerate}
+
+The discrete-time STFT is defined in (\ref{stft}) where $x[n]$ is the discrete signal and $w[n]$ is the window function.
+
+\begin{equation}
+	\label{stft}
+	\mathbf{STFT}\{x[n]\}(m,\omega) \equiv X(m,\omega) = \sum_{n=-\infty}^{\infty}x[n]w[n-m]e^{-j \omega n}
+\end{equation}
+
+The Mel scale is defined in (\ref{melScale}) where $f$ is the frequency in hertz and $m$ is mels.\cite{Oshaughnessy1990}
+
+\begin{equation}
+\label{melScale}
+m = 2595 log_{10}(1 + \frac{f}{700})
+\end{equation}
 
@@
 	\caption{A high level overview of the processing stages that a speech sample goes through.}
-	\label{fig:high_level_dataflow_diagram}
+	\label{highLevelDataFlowDiagram}
 \end{figure}
@@
 
-\begin{figure*}[b!]
+\begin{figure*}[h!]
 	\centering
@@
 	\includegraphics[width=\textwidth]{neural_network_architecture.png}
-	\caption{Architecture for the speech emotion recognition network including the input layer and seven processing layers.}
-	\label{fig:neural_network_architecture}
+	\caption{Architecture for the speech emotion recognition network including the input layer and eight processing layers.}
+	\label{neuralNetworkArchitecture}
 \end{figure*}
@@
 After all of the databases were processed this way, the final combined database was fed into a neural network for training, validation, and testing.
-We constructed a seven layer neural network consisting of four convolutional layers and three dense layers.
-The network architecture is shown in figure \ref{fig:neural_network_architecture}.
+We constructed a eight layer neural network consisting of four convolutional layers and three dense layers.
+The network architecture is shown in figure \ref{neuralNetworkArchitecture}.
 The training process is described below.
@@
 	\item We split the combined database into 80\% training, 10\% validation, and 10\% testing.
-	\item We applied dropout \cite{Srivastava2014} to the dense layers and batch normalization \cite{Ioffe2015} to the convolutional layers to deal with overfitting.
+	\item We applied dropout to the dense layers and batch normalization to the convolutional layers to deal with overfitting. \cite{Srivastava2014}, \cite{Ioffe2015}
 	\item We updated the model's hyperparameters based on the validation loss and accuracy to improve the model's accuracy and ability to generalize.
@@
 
-Table \ref{tab:model_hyperparams} describes the network hyperparameters and table \ref{tab:train_hyperparams} describes the training hyperparameters.
+Table \ref{modelHyperparams} describes the network hyperparameters and table \ref{trainHyperparams} describes the training hyperparameters.
 
-\begin{table}[b]
+\begin{table}[h!]
+	\caption{The model's hyperparameters.}
+	\label{modelHyperparams}
 	\resizebox{\linewidth}{!}{%
@@
 	}
-	\caption{The model's hyperparameters.}
-	\label{tab:model_hyperparams}
 \end{table}
 
-\begin{table}[]
+\begin{table}[h!]
+	\caption{The training hyperparameters.}
+	\label{trainHyperparams}
 	\resizebox{\linewidth}{!}{%
@@
 	}
-	\caption{The training hyperparameters.}
-	\label{tab:train_hyperparams}
 \end{table}
@@
 
-After training the model, the model was evaluated on the testing set to get the final accuracy of the model.
+After training the model, it was evaluated on the testing set to get the final accuracy.
 The final accuracy achieved was 52.57\%.
-Confusion matrices for each emotion are shown in Figure \ref{fig:confusion_matrix}.
+Confusion matrices for each emotion are shown in Figure \ref{confusionMatrix}.
 
-\begin{figure}[]
+\begin{figure*}[h!]
 	\centering
@@
 	\caption{Confusion matrices for each emotion.}
-	\label{fig:confusion_matrix}
-\end{figure}
+	\label{confusionMatrix}
+\end{figure*}
 
@@
 
-We created a neural network model that achieved a classification accuracy of 52.57\% on the problem of multi-label emotion classification.
-This model can successfully classify multiple emotions in a sample and while this is a good result, we can do better.
+We created a neural network model that achieved a 52.57\% accuracy on the problem of multi-label speech emotion classification.
+This is better than the state-of-the-art result of 52.14\% achieved in the single-label case.
+\cite{Zhao2019} So we conclude that it is possible to achieve a reasonable accuracy on the problem of multi-label speech emotion recognition by applying deep learning to the problem.
+
+In analyzing the confusion matrices, we see that the model predicts an emotion is absent in most of the samples with "surprise" being the exception.
+We suspect that "surprise" being an exception is due to the class imbalance that is shown in figure \ref{combinedDb}.
+The "surprise" class was the least represented class out of the seven classes and this makes the model predict it more due to the use of class weights.
+Class weights bias the model towards underrepresented classes in an attempt to balance classes but this seems to have affected the model's ability to learn.
 
@@
 
-One limitation of this work was that we only accounted for seven emotions but we know that there are more emotions than this. \cite{} Another limitations was that 
-Here we present future improvements of our work 
+One limitation of this work is that we only accounted for seven emotions but recent research has suggested that there are more emotions such as boredom, shame, and triumph.
+\cite{Cordaro2018} However, one issue with expanding the set of considered emotions is the lack of databases with the labeled emotion.
+
+Another limitation of this work is that all samples are spoken in English so the model is biased towards Anglophones.
+In theory, the basic emotions are universal across languages and cultures so incorporating databases spoken in different languages, such as the Emo-DB database, would help the model generalize across languages.\cite{Burkhardt2005}
+
+The following list is a suggestion that future work could pursue:
 \begin{itemize}
-	\item Improving the recognition accuracy by using more sophisticated neural network architectures such as LSTMs or using more databases.
-	\item Incorporating the phase data from the STFT and testing to see if that improves the accuracy.
-	\item Testing a binary relevant approach to this multi-label problem and comparing it to this model.
+	\item Using more sophisticated neural network architectures such as LSTMs or using more databases.
+	\item Incorporating the phase data from the STFT.
+	\item Testing a binary relevance approach to this multi-label problem.
+	\item Replacing the use of STFT with a wavelet transform.
 \end{itemize}
@@
 
-Overall, this paper presented a convolutional neural network model that achieves a 52.57\% accuracy on multi-class, multi-label speech emotion recognition using four combined databases.
-We obtained this result by transforming raw speech samples into log-Mel spectrograms using STFT and the Mel scale.
-These log-Mel spectrograms are then fed into a eight layer neural network for classification for training.
-When we have completed training the neural network, the model was evaluated and achieved as 52.57\% accuracy.
-While this result is a promising start, we suggest improvements that future work can build upon to improve the accuracy.
+Overall, this paper presented a 2D CNN model that achieved an accuracy of 52.57\% on the problem of multi-label speech emotion recognition using four combined databases.
+We obtained this result by transforming raw speech samples into log-Mel spectrograms using the STFT and the Mel scale.
+The log-Mel spectrograms are then fed into an eight layer neural network for classification.
+While this result is a promising start, we suggest improvements that future work can build upon to improve the accuracy of the model and to expand the scope of emotions considered.
 
 \section*{Acknowledgment}
-Acknowledge PURE award.
+This work was funded by the Program for Undergraduate Research Experience (PURE) award granted by the University of Calgary.
 

### 2019-09-04 research/paper/paper.tex
Update paper with revisions

@@
 \section{Introduction}
-
-\subsection{Literature Review}
-
 The field of affective computing studies the development of systems that can recognize, interpret, process, and simulate human affects.
@@
 \end{itemize}
+
+\subsection{Literature Review}
+
 Feature engineering is a method of manually extracting desired parameters from the data for use in  recognition.
@@
 
-\subsection{Summary}
+\subsection{Proposed Approach}
 
@@
 These log-Mel spectrograms were then fed into an eight-layer neural network consisting of four convolutional layers and four dense layers.
-After training the neural network, the model achieves a 52.57\% test accuracy.
 
@@
 \section{Methodology}
+
+We approach the problem of SER by preprocessing samples from four databases and then feeding those samples into a CNN.
+We preprocess the speech samples into log-Mel spectrogram to help the neural network extract features relevant to emotion recognition.
+This choice is based on previous work such as \cite{Engel2019} and \cite{Badshah2017}.
+The choice of using a CNN is based on previous work where Balakrishnan et al.
+\cite{Balakrishnan2017} found that CNN-based models have superior performance compared to RNNs.
+Another justification for using CNNs is due treating the log-Mel spectrograms as images where CNNs have been shown to perform well on.\cite{Krizhevsky2012}
+
+A high level overview of the data flow is shown in figure \ref{highLevelDataFlowDiagram} where samples flow through the four stages of processing.
+Both the preprocessing steps and the CNN were implemented in Python using the Librosa \cite{McFee2015} and Keras \cite{Chollet2015} library respectively.
 
@@
 
-Each sample from a database would flow through the same preprocessing steps to maintain consistency across all samples.
-A high level overview of the preprocessing steps is shown in figure \ref{highLevelDataFlowDiagram} and the detailed steps are described below.
+Each sample from a database would flow through the same preprocessing steps to maintain consistency across all samples. The detailed steps are described below.
 \begin{enumerate}
@@
 
-For the STFT, we used a window size of 3072 with a 75\% overlap. For the Mel spectrogram, we set the minimum frequency to 20 Hz and maximum frequency to 12 kHz with 200 Mel bins.
+For the STFT, we used a window size of 3072 with a 75\% overlap.
+This choice is based on the work of Zhao et al.
+\cite{Zhao2019} where they also use 75\% overlap but with a smaller window size (2048) and they achieved excellent results.
+For the Mel spectrogram, we set the minimum frequency to 20 Hz and maximum frequency to 12 kHz with 200 Mel bins.
+The frequency range was chosen after experimenting with various ranges and selecting the range that resulted in visually clean spectrograms.
+The number of Mel bins was also chosen after experimentation as too few bins resulted in poor temporal resolution while too many bins resulted in poor frequency resolution.
+This is due to the Gabor limit.
 
@@
 
+\bgroup
+\def\arraystretch{2}
 \begin{table}[h!]
+	\centering
 	\caption{The model's hyperparameters.}
 	\label{modelHyperparams}
-	\resizebox{\linewidth}{!}{%
+	{%
 		\begin{tabular}{|l|r|}
@@
 \end{table}
+\egroup
 
+\bgroup
+\def\arraystretch{2}
 \begin{table}[h!]
+	\centering
 	\caption{The training hyperparameters.}
 	\label{trainHyperparams}
-	\resizebox{\linewidth}{!}{%
+	{%
 		\begin{tabular}{|l|r|}
@@
 \end{table}
+\egroup
 
-\section{Results}
+The final model is evaluated on the testing set using two measures: categorical cross-entropy and accuracy.
+Categorical cross-entropy as defined in (\ref{cce}) where $N$ is the number of samples, $k$ is the number of classes, $t_{i,j}$ is 1 if the sample $i$ is in class $j$ and 0 otherwise, and $p_{i,j}$ is the model's predicted probability that sample $i$ is in class $j$.
+\begin{equation}
+\label{cce}
+CCE = -\frac{1}{N}\sum_{i=1}^{N}\sum_{j=1}^{k}t_{i,j}log(p_{i,j})
+\end{equation}
+The choice to use categorical cross-entropy is based on its ease of use and how prevalent it is used in mutli-label problems.
+
+The second measure, accuracy, is calculated from the confusion matrices shown in \ref{confusionMatrix}.
+The confusion matrices are calculated by comparing the true labels from the predicted labels for each emotion.
+There is a confusion matrix for each emotion due to the multi-label problem considering each emotion as independent.
+We define accuracy as the unweighted average accuracy of all emotions and this is mathematically described in (\ref{aveacc}) where $n$ is the total number of emotions (7 in this case), $TP_{i}$ is the number of true positives for emotion $i$, $TN_{i}$ is the number of true negatives for emotion $i$, and total is the total number of predictions.
+
+\begin{equation}
+\label{aveacc}
+Accuracy = \frac{\sum_{i=1}^{n}\frac{TP_{i} + TN_{i}}{Total_{i}}}{n}
+\end{equation}
+
+We chose this definition of accuracy so that we can compare our results to other papers that also use a similar measure.
+
+\section{Results and Discussion}
 
@@
 
-\section{Discussion}
-
 We created a neural network model that achieved a 52.57\% accuracy on the problem of multi-label speech emotion classification.
@@
 
-\subsection{Limitations and Future Work}
+\section{Conclusion and Future Work}
+
+Overall, this paper presented a 2D CNN model that achieved an accuracy of 52.57\% on the problem of multi-label speech emotion recognition using four combined databases.
+We obtained this result by transforming raw speech samples into log-Mel spectrograms using the STFT and the Mel scale.
+The log-Mel spectrograms are then fed into an eight layer neural network for classification.
+While this result is a promising start, we suggest improvements that future work can build upon to improve the accuracy of the model and to expand the scope of emotions considered.
 
@@
 \end{itemize}
-
-\subsection{Conclusion}
-
-Overall, this paper presented a 2D CNN model that achieved an accuracy of 52.57\% on the problem of multi-label speech emotion recognition using four combined databases.
-We obtained this result by transforming raw speech samples into log-Mel spectrograms using the STFT and the Mel scale.
-The log-Mel spectrograms are then fed into an eight layer neural network for classification.
-While this result is a promising start, we suggest improvements that future work can build upon to improve the accuracy of the model and to expand the scope of emotions considered.
+Rana et al. \cite{Rana2016} has also shown that SER systems can be more robust by introducing noise into the samples which is another promising future direction.
 

### 2019-09-04 src/database_processor/display_cm.py

@@ -1,0 +1,85 @@
+"""
+This file holds a specialized function to display a confusion matrix for use in the paper.
+"""
+
+import matplotlib.pyplot as plt
+import numpy as np
+
+from src import em_constants as emc
+
+CM = np.array([
+        [[1.00, 0.00],
+         [0.94, 0.06]],
+        [[1.00, 0.00],
+         [0.97, 0.03]],
+        [[1.00, 0.00],
+         [0.87, 0.13]],
+        [[1.00, 0.00],
+         [0.79, 0.21]],
+        [[1.00, 0.00],
+         [0.97, 0.03]],
+        [[1.00, 0.00],
+         [0.86, 0.14]],
+        [[1.00, 0.00],
+         [0.53, 0.47]],
+    ])
+
+
+
+def visualize_confusion_matrix():
+    """
+    Visualizes the test set confusion matrix.
+    """
+    plt.rcParams.update({'font.size': 21})
+
+    # Display the confusion matrix for each emotion
+    labels = ["Absent", "Present"]
+
+    for em_index in range(0, emc.NUM_EMOTIONS):
+        # Normalize the confusion matrix
+        em_cm = CM[em_index]
+        em_cm = em_cm.astype('float') / em_cm.sum(axis=1)[:, np.newaxis]
+
+        # Plot the confusion matrix
+        fig, ax = plt.subplots()
+        im = ax.imshow(em_cm, interpolation='nearest',
+                       cmap=plt.get_cmap("Blues"))
+        ax.figure.colorbar(im, ax=ax)
+
+        # Set the x and y axis labels
+        em_label = emc.INVERT_EMOTION_MAP[em_index].title()
+        ax.set(xticks=np.arange(em_cm.shape[1]),
+               yticks=np.arange(em_cm.shape[0]),
+               # ... and label them with the respective list entries
+               xticklabels=labels, yticklabels=labels,
+               ylabel='True label',
+               xlabel='Predicted label')
+
+        # Rotate the tick labels and set their alignment.
+        plt.setp(ax.get_xticklabels(), rotation=45, ha="right",
+                 rotation_mode="anchor")
+        ax.set_ylim(len(em_cm) - 0.5, -0.5)
+        ax.set_title('Normalized Confusion Matrix for {}'.format(em_label),
+                     pad=20)
+
+        # Loop over data dimensions and create text annotations.
+        fmt = '.2f'
+        thresh = em_cm.max() / 2.
+        for i in range(em_cm.shape[0]):
[…]

### 2019-09-05 research/paper/paper.tex
Update paper draft

@@
 We created a combined database consisting of four speech emotion-labeled databases, and we used it to train a 2D convolutional neural network to determine if the model could recognize multiple emotions in a speech sample.
-The model was able to classify the samples with an accuracy of 52.57\%.
-This shows that it is possible to apply machine learning to the problem of multi-label speech emotion recognition and to achieve a reasonable accuracy.\footnote{\url{https://github.com/Brian-Pho/RVST598_Speech-Emotion-Recognition}}
+The model was able to classify the samples with an accuracy of 57.64\%.
+This shows that it is possible to apply machine learning to the problem of multi-label speech emotion recognition and to achieve a reasonable accuracy.\footnote{Code for this paper is available on Github at: \url{https://github.com/Brian-Pho/RVST598_Speech-Emotion-Recognition}}
 \end{abstract}
@@
 
-Current state-of-the-art speech emotion recognition has achieved an accuracy of 52.14\% for the case of speaker-independent, single-labeled emotions on the IEMOCAP database.\cite{Zhao2019} This result was achieved using a 2D CNN long short-term memory (LSTM) model where the audio waveform was converted into a log-Mel spectrogram and then fed into the model.
+Current state-of-the-art SER has achieved an accuracy of 52.14\% for the case of speaker-independent, single-labeled emotions on the IEMOCAP database.\cite{Zhao2019} This result was achieved using a 2D CNN long short-term memory (LSTM) model where the audio waveform was converted into a log-Mel spectrogram and then fed into the model.
 
@@
 Kim et al.
-\cite{Kim2018a} is an exception as they consider the multi-label case of emotion recognition but they did not use machine learning and they used visual data on top of audio data.
-This paper extends upon the state-of-the-art by considering the problem of multi-label speech emotion classification using four databases to train and test the neural network.
+\cite{Kim2018a} is an exception as they consider the multi-label case of SER but they did not use machine learning and they used visual data on top of audio data.
+This paper extends upon current literature by considering the problem of multi-label SER using four databases to train and test a CNN.
 
@@
 	\item Section II details the methodology we use such as the how we combined the four databases and describes the architecture of the neural network.
-	\item Section III describes the results from testing the neural network.
-	\item Section IV discusses the results in the context of the field and how the results can be improved upon.
+	\item Section III describes and discusses the results from testing the neural network.
+	\item Section IV concludes this paper and provides suggestions for future directions.
 \end{itemize}
@@
 This choice is based on previous work such as \cite{Engel2019} and \cite{Badshah2017}.
-The choice of using a CNN is based on previous work where Balakrishnan et al.
+The choice of using a CNN is also based on previous work where Balakrishnan et al.
 \cite{Balakrishnan2017} found that CNN-based models have superior performance compared to RNNs.
-Another justification for using CNNs is due treating the log-Mel spectrograms as images where CNNs have been shown to perform well on.\cite{Krizhevsky2012}
+Another justification for using CNNs is due treating the log-Mel spectrograms as images which CNNs have been shown to perform well on.\cite{Krizhevsky2012}
 
-A high level overview of the data flow is shown in figure \ref{highLevelDataFlowDiagram} where samples flow through the four stages of processing.
+A high level overview of the data flow is shown in figure \ref{highLevelDataFlowDiagram} where samples flow through four stages of processing.
+The stages are detailed further into the methodology.
 Both the preprocessing steps and the CNN were implemented in Python using the Librosa \cite{McFee2015} and Keras \cite{Chollet2015} library respectively.
@@
 The number of Mel bins was also chosen after experimentation as too few bins resulted in poor temporal resolution while too many bins resulted in poor frequency resolution.
-This is due to the Gabor limit.
+This is due to the Gabor limit which describes the trade-off between temporal and frequency resolution.
 
@@
 	\includegraphics[width=\linewidth]{high_level_dataflow_diagram.png}
-	\caption{A high level overview of the processing stages that a speech sample goes through.}
+	\caption{High level overview of the processing stages that a speech sample goes through.}
 	\label{highLevelDataFlowDiagram}
@@
 After all of the databases were processed this way, the final combined database was fed into a neural network for training, validation, and testing.
-We constructed a eight layer neural network consisting of four convolutional layers and three dense layers.
+We constructed an eight layer neural network consisting of four convolutional layers and three dense layers.
 The network architecture is shown in figure \ref{neuralNetworkArchitecture}.
@@
 
-The final model is evaluated on the testing set using two measures: categorical cross-entropy and accuracy.
-Categorical cross-entropy as defined in (\ref{cce}) where $N$ is the number of samples, $k$ is the number of classes, $t_{i,j}$ is 1 if the sample $i$ is in class $j$ and 0 otherwise, and $p_{i,j}$ is the model's predicted probability that sample $i$ is in class $j$.
-\begin{equation}
-\label{cce}
-CCE = -\frac{1}{N}\sum_{i=1}^{N}\sum_{j=1}^{k}t_{i,j}log(p_{i,j})
-\end{equation}
-The choice to use categorical cross-entropy is based on its ease of use and how prevalent it is used in mutli-label problems.
+The final model was evaluated on the testing set by using average accuracy.
+The average accuracy is calculated from the confusion matrices shown in figure \ref{confusionMatrix}.
+The confusion matrices are calculated by comparing the true labels to the predicted labels for each emotion.
+There is a confusion matrix for each emotion due to the multi-label problem considering each emotion as independent.
 
-The second measure, accuracy, is calculated from the confusion matrices shown in \ref{confusionMatrix}.
-The confusion matrices are calculated by comparing the true labels from the predicted labels for each emotion.
-There is a confusion matrix for each emotion due to the multi-label problem considering each emotion as independent.
-We define accuracy as the unweighted average accuracy of all emotions and this is mathematically described in (\ref{aveacc}) where $n$ is the total number of emotions (7 in this case), $TP_{i}$ is the number of true positives for emotion $i$, $TN_{i}$ is the number of true negatives for emotion $i$, and total is the total number of predictions.
+We define accuracy as the unweighted average accuracy of all emotions and this is mathematically described in (\ref{aveacc}) where $n$ is the total number of emotions (7 in our case), $TP_{i}$ is the number of true positives for emotion $i$, $TN_{i}$ is the number of true negatives for emotion $i$, and $Total_{i}$ is the total number of predictions for emotion $i$.
 
@@
 
-We chose this definition of accuracy so that we can compare our results to other papers that also use a similar measure.
+In summary, we calculate the accuracy per emotion and then average over these accuracies to obtain the final accuracy.
+We chose this definition of accuracy so that we can compare our results to other papers that also use a similar measure of accuracy that is calculated from the confusion matrices.
 
@@
 After training the model, it was evaluated on the testing set to get the final accuracy.
-The final accuracy achieved was 52.57\%.
-Confusion matrices for each emotion are shown in Figure \ref{confusionMatrix}.
+The final accuracy achieved was 57.64\%.
+Confusion matrices for each emotion are shown in Figure \ref{confusionMatrix} and comparisons to the current literature is shown in table \ref{litacccompare}.
 
@@
 
-We created a neural network model that achieved a 52.57\% accuracy on the problem of multi-label speech emotion classification.
-This is better than the state-of-the-art result of 52.14\% achieved in the single-label case.
-\cite{Zhao2019} So we conclude that it is possible to achieve a reasonable accuracy on the problem of multi-label speech emotion recognition by applying deep learning to the problem.
+\bgroup
+\def\arraystretch{2}
+\begin{table}[]
+	\centering
+	\caption{Comparison of SER accuracy in literature.}
+	\label{litacccompare}
+	\begin{tabular}{|l|r|r|}
+		\hline
+		\multicolumn{1}{|c|}{\textbf{Research Work}} & \multicolumn{1}{c|}{\textbf{Method}} & \multicolumn{1}{c|}{\textbf{Accuracy (\%)}} \\ \hline
+		\rowcolor[HTML]{F6F8FA}
+		Zhao et al. \cite{Zhao2019} & CNN + LSTM & 52.1 \\ \hline
+		Our Work & CNN & 57.6 \\ \hline
+		\rowcolor[HTML]{F6F8FA}
+		Etienne et al. \cite{Etienne2018} & CNN + LSTM & 61.7 \\ \hline
+		Zhang et al. \cite{Zhang2019} & CNN & 63.9 \\ \hline
+		\rowcolor[HTML]{F6F8FA}
+		Fayek et al. \cite{Fayek2017} & CNN & 64.8 \\ \hline
+		Yenigalla et al. \cite{Yenigalla2018} & CNN & 73.9 \\ \hline
+		\rowcolor[HTML]{F6F8FA}
+		Badshah et al. \cite {Badshah2019} & CNN & 84.3 \\ \hline
+	\end{tabular}
+\end{table}
+\egroup
+
+The results achieved in this paper do not reach the state-of-the-art accuracy achieved for the single-label case but this is the first attempt on the harder case of mutli-label using more data.
+So we conclude that it is possible to achieve a reasonable accuracy on the problem of multi-label speech emotion recognition by applying deep learning to the problem.
 
@@
 
-Overall, this paper presented a 2D CNN model that achieved an accuracy of 52.57\% on the problem of multi-label speech emotion recognition using four combined databases.
+Overall, this paper presented a 2D CNN model that achieved an accuracy of 57.64\% on the problem of multi-label speech emotion recognition using four combined databases.
 We obtained this result by transforming raw speech samples into log-Mel spectrograms using the STFT and the Mel scale.

### 2019-09-06 src/database_processor/display_cm.py
Update diagrams and shorten STFT equation

@@ -25,7 +25,6 @@ CM = np.array([
     ])
 
 
-
 def visualize_confusion_matrix():
     """
     Visualizes the test set confusion matrix.

### 2019-09-07 research/paper/paper_body.tex
Change paper template to ACM SAC 2020

--- 2019-09-05 research/paper/paper.tex
@@
-\documentclass[conference]{IEEEtran}
-\IEEEoverridecommandlockouts
+\section{Introduction}
 
-\usepackage{cite}
-\usepackage{amsmath,amssymb,amsfonts}
-\usepackage{algorithmic}
-\usepackage{graphicx}
-\usepackage{dblfloatfix}
-\usepackage{hyperref}
-\usepackage{caption}
-\usepackage{pifont}
-\usepackage{textcomp}
-\usepackage{multirow}
-\usepackage[table,xcdraw]{xcolor}
-\usepackage[strings]{underscore}
-
-\graphicspath{{figures/}}
-\newcommand{\cmark}{\ding{51}}
-\newcommand{\xmark}{\ding{55}}
-\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08emT\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
-
-\begin{document}
-
-\title{Multi-Label Speech Emotion Recognition Using 2D Convolutional Neural Networks}
-
-\author{\IEEEauthorblockN{Brian Pho, Thomas Truong, Svetlana Yanushkevich}
-\IEEEauthorblockA{\textit{Biometric Technologies Laboratory, Department of Electrical and Computer Engineering}\\
-University of Calgary, Canada\\
-\{brian.pho, thomas.truong, syanshk\}@ucalgary.ca}}
-
-\maketitle
-
-\begin{abstract}
-Current speech emotion recognition systems often overlook the multi-label data that comes with databases.
-In this paper, we address the problem of whether machine learning can be used to detect multiple emotions in speech.
-We created a combined database consisting of four speech emotion-labeled databases, and we used it to train a 2D convolutional neural network to determine if the model could recognize multiple emotions in a speech sample.
-The model was able to classify the samples with an accuracy of 57.64\%.
-This shows that it is possible to apply machine learning to the problem of multi-label speech emotion recognition and to achieve a reasonable accuracy.\footnote{Code for this paper is available on Github at: \url{https://github.com/Brian-Pho/RVST598_Speech-Emotion-Recognition}}
-\end{abstract}
-
-\begin{IEEEkeywords}
-multi-label, speech, emotion, recognition, neural networks
-\end{IEEEkeywords}
-
-\section{Introduction}
 The field of affective computing studies the development of systems that can recognize, interpret, process, and simulate human affects.
@@
 
-\bgroup
-\def\arraystretch{2}
-\begin{table*}
-	\resizebox{\textwidth}{!}{%
-		\begin{tabular}{|l|c|c|c|c|c|c|c|c|c|c|c|c|c|}
-			\hline
-			\multicolumn{1}{|c|}{} & \multicolumn{13}{c|}{\textbf{Emotions}} \\ \cline{2-14} 
-			\multicolumn{1}{|c|}{\multirow{-2}{*}{\textbf{Databases}}} & \textbf{Neutral} & \textbf{Anger} & \textbf{Disgust} & \textbf{Fear} & \textbf{Happy} & \textbf{Sad} & \textbf{Surprise} & \textbf{Calm} & \textbf{Excitement} & \textbf{Frustration} & \textbf{Amused} & \textbf{Sleepy} & \textbf{Bored} \\ \hline
-			\rowcolor[HTML]{F6F8FA} 
-			\textbf{IEMOCAP} & \cmark & \cmark & \cmark & \cmark & \cmark & \cmark & \cmark & \xmark & \cmark & \cmark & \xmark & \xmark & \xmark \\ \hline
-			\textbf{CREMA-D} & \cmark & \cmark & \cmark & \cmark & \cmark & \cmark & \xmark & \xmark & \xmark & \xmark & \xmark & \xmark & \xmark \\ \hline
-			\rowcolor[HTML]{F6F8FA} 
-			\textbf{TESS} & \cmark & \cmark & \cmark & \cmark & \cmark & \cmark & \cmark & \xmark & \xmark & \xmark & \xmark & \xmark & \xmark \\ \hline
-			\textbf{RAVDESS} & \cmark & \cmark & \cmark & \cmark & \cmark & \cmark & \cmark & \cmark & \xmark & \xmark & \xmark & \xmark & \xmark \\ \hline
-			\rowcolor[HTML]{F6F8FA} 
-			\textbf{MSP-IMPROV} & \cmark & \cmark & \cmark & \cmark & \cmark & \cmark & \cmark & \xmark & \xmark & \xmark & \xmark & \xmark & \xmark \\ \hline
-			\textbf{SAVEE} & \cmark & \cmark & \cmark & \cmark & \cmark & \cmark & \cmark & \xmark & \xmark & \xmark & \xmark & \xmark & \xmark \\ \hline
-			\rowcolor[HTML]{F6F8FA} 
-			\textbf{Emo-DB} & \cmark & \cmark & \cmark & \cmark & \cmark & \cmark & \xmark & \xmark & \xmark & \xmark & \xmark & \xmark & \cmark \\ \hline
-			\textbf{EmoV-DB} & \cmark & \cmark & \cmark & \xmark & \xmark & \xmark & \xmark & \xmark & \xmark & \xmark & \cmark & \cmark & \xmark \\ \hline
-		\end{tabular}%
-	}
-	\caption{Comparison of databases with their labeled emotions.}
-	\label{dbEmoTable}
-\end{table*}
-\egroup
-
 \section{Methodology}
@@
 \subsection{Preprocessing}
+
+\begin{table*}
+	\centering
+	\caption{Comparison of databases with their labeled emotions.}
+	\label{dbEmoTable}
+	\begin{tabular}{@{}lccccccccccccc@{}}
+		\toprule
+		Databases & \multicolumn{13}{c}{Emotions} \\ 
+		\cmidrule{2-14}
+		& Neutral & Anger & Disgust & Fear & Happy & Sad & Surprise & Calm & Excitement & Frustration & Amused & Sleepy & Bored \\ 
+		\midrule
+		IEMOCAP & x & x & x & x & x & x & x &   & x & x &   &   &   \\
+		CREMA-D & x & x & x & x & x & x &   &   &   &   &   &   &   \\
+		TESS & x & x & x & x & x & x & x &   &   &   &   &   &   \\
+		RAVDESS & x & x & x & x & x & x & x & x &   &   &   &   &   \\
+		MSP-IMPROV & x & x & x & x & x & x & x &   &   &   &   &   &   \\
+		SAVEE & x & x & x & x & x & x & x &   &   &   &   &   &   \\
+		Emo-DB & x & x & x & x & x & x &   &   &   &   &   &   & x \\
+		EmoV-DB & x & x & x &   &   &   &   &   &   &   & x & x &   \\
+		\bottomrule
+	\end{tabular}%
+\end{table*}
 
@@
 
-\begin{figure*}[t]
+\begin{figure*}
 	\centering
@@
 	\label{stft}
-	\mathbf{STFT}\{x[n]\}(m,\omega) \equiv X(m,\omega) = \sum_{n=-\infty}^{\infty}x[n]w[n-m]e^{-j \omega n}
+	\mathbf{STFT}\{x[n]\}(m,\omega) = \sum_{n=-\infty}^{\infty}x[n]w[n-m]e^{-j \omega n}
 \end{equation}
@@
 
-\begin{figure}[]
+\begin{figure}
 	\centering
@@
 
-\begin{figure*}[h!]
+\begin{figure*}
 	\centering
@@
 
-\bgroup
-\def\arraystretch{2}
-\begin{table}[h!]
+\begin{table}
 	\centering
@@
 	\label{modelHyperparams}
-	{%
-		\begin{tabular}{|l|r|}
-			\hline
-			\textbf{Hyperparameter} & \textbf{Value} \\ \hline
-			\rowcolor[HTML]{F6F8FA} 
-			Input Dimensions & 278w x 200h x 1d \\ \hline
-			Optimization Algorithm & Adam \\ \hline
-			\rowcolor[HTML]{F6F8FA} 
-			Loss Measure & Binary Cross-entropy \\ \hline
-			Accuracy Metric & Categorical Cross-entropy \\ \hline
-			\rowcolor[HTML]{F6F8FA} 
-			Activation Function & Sigmoid \\ \hline
-		\end{tabular}%
-	}
+	\begin{tabular}{@{}lr@{}}
+		\toprule
+		Hyperparameter & Value \\
+		\midrule
+		Input Dimensions & 278w x 200h x 1d \\
+		Optimization Algorithm & Adam \\
+		Loss Measure & Binary Cross-entropy \\
+		Accuracy Metric & Categorical Cross-entropy \\
+		Activation Function & Sigmoid \\
+		\bottomrule
+	\end{tabular}
 \end{table}
-\egroup
 
-\bgroup
-\def\arraystretch{2}
-\begin{table}[h!]
+\begin{table}
 	\centering
@@
 	\label{trainHyperparams}
-	{%
-		\begin{tabular}{|l|r|}
-			\hline
-			\textbf{Hyperparameter} & \textbf{Value} \\ \hline
-			\rowcolor[HTML]{F6F8FA} 
-			Epochs & 20 \\ \hline
-			Batch Size & 32 \\ \hline
-			\rowcolor[HTML]{F6F8FA} 
-			Training Set & 17,341 \\ \hline
-			Validation Set & 2,167 \\ \hline
-			\rowcolor[HTML]{F6F8FA} 
-			Testing Set & 2,167 \\ \hline
-		\end{tabular}%
-	}
+	\begin{tabular}{@{}lr@{}}
+		\toprule
+		Hyperparameter & Value \\
+		\midrule
+		Epochs & 20 \\
+		Batch Size & 32 \\
+		Training Set & 17,341 \\
+		Validation Set & 2,167 \\
+		Testing Set & 2,167 \\
+		\bottomrule
+	\end{tabular}
 \end{table}
-\egroup
 
@@
 
-\bgroup
-\def\arraystretch{2}
-\begin{table}[]
+\begin{table}
 	\centering
@@
 	\label{litacccompare}
-	\begin{tabular}{|l|r|r|}
-		\hline
-		\multicolumn{1}{|c|}{\textbf{Research Work}} & \multicolumn{1}{c|}{\textbf{Method}} & \multicolumn{1}{c|}{\textbf{Accuracy (\%)}} \\ \hline
-		\rowcolor[HTML]{F6F8FA}
-		Zhao et al. \cite{Zhao2019} & CNN + LSTM & 52.1 \\ \hline
-		Our Work & CNN & 57.6 \\ \hline
-		\rowcolor[HTML]{F6F8FA}
-		Etienne et al. \cite{Etienne2018} & CNN + LSTM & 61.7 \\ \hline
-		Zhang et al. \cite{Zhang2019} & CNN & 63.9 \\ \hline
-		\rowcolor[HTML]{F6F8FA}
-		Fayek et al. \cite{Fayek2017} & CNN & 64.8 \\ \hline
-		Yenigalla et al. \cite{Yenigalla2018} & CNN & 73.9 \\ \hline
-		\rowcolor[HTML]{F6F8FA}
-		Badshah et al. \cite {Badshah2019} & CNN & 84.3 \\ \hline
+	\begin{tabular}{@{}lrr@{}}
+		\toprule
+		Research Work & Method & Accuracy (\%) \\
+		\midrule
+		Zhao et al. \cite{Zhao2019} & CNN + LSTM & 52.1 \\
+		Our Work & CNN & 57.6 \\
+		Etienne et al. \cite{Etienne2018} & CNN + LSTM & 61.7 \\
+		Zhang et al. \cite{Zhang2019} & CNN & 63.9 \\
+		Fayek et al. \cite{Fayek2017} & CNN & 64.8 \\
+		Yenigalla et al. \cite{Yenigalla2018} & CNN & 73.9 \\
+		Badshah et al. \cite {Badshah2019} & CNN & 84.3 \\
+		\bottomrule
 	\end{tabular}
 \end{table}
-\egroup
 
@@
 
-\section*{Acknowledgment}
+\begin{acks}
 This work was funded by the Program for Undergraduate Research Experience (PURE) award granted by the University of Calgary.
-
-\bibliographystyle{IEEEtran}
-\bibliography{IEEEabrv,references}
-
-\end{document}
+\end{acks}

### 2019-09-07 research/paper/paper_header.tex
Change paper template to ACM SAC 2020

\documentclass[sigconf]{acmart}

\usepackage{booktabs} % For formal tables
\usepackage{graphicx}
\graphicspath{{figures/}}

% Copyright
%\setcopyright{none}
\setcopyright{acmcopyright}
%\setcopyright{acmlicensed}
%\setcopyright{rightsretained}
%\setcopyright{usgov}
%\setcopyright{usgovmixed}
%\setcopyright{cagov}
%\setcopyright{cagovmixed}

% DOI
\acmDOI{xx.xxx/xxx_x}

% ISBN
\acmISBN{978-1-4503-6866-7/20/03}

%Conference
\acmConference[SAC'20]{ACM SAC Conference}{March 30-April 3, 2020}{Brno, Czech Republic} 
\acmYear{2020}
\copyrightyear{2020}

\acmArticle{4}
\acmPrice{15.00}

% These commands are optional
%\acmBooktitle{Transactions of the ACM Woodstock conference}
%\editor{Jennifer B. Sartor}
%\editor{Theo D'Hondt}
%\editor{Wolfgang De Meuter}

\begin{document}
\title{Multi-Label Speech Emotion Recognition Using 2D Convolutional Neural Networks}
\subtitle{Using machine learning to identify multiple emotions in speech}

\author{Brian Pho}
\affiliation{%
	\institution{University of Calgary}
	\streetaddress{2500 University Drive NW}
	\city{Calgary} 
	\state{Alberta} 
	\postcode{T2N 1N4}
}
\email{brian.pho@ucalgary.ca}

\author{Thomas Truong}
\affiliation{%
	\institution{University of Calgary}
	\streetaddress{2500 University Drive NW}
	\city{Calgary} 
	\state{Alberta} 
	\postcode{T2N 1N4}
}
\email{thomas.truong@ucalgary.ca}

\author{Svetlana Yanushkevich}
\affiliation{%
	\institution{University of Calgary}
	\streetaddress{2500 University Drive NW}
	\city{Calgary} 
	\state{Alberta} 
	\postcode{T2N 1N4}
}
\email{syanshk@ucalgary.ca}

% The default list of authors is too long for headers}
\renewcommand{\shortauthors}{B. Pho et al.}

\begin{abstract}
Current speech emotion recognition systems often overlook the multi-label data that comes with databases. In this paper, we address the problem of whether machine learning can be used to detect multiple emotions in speech. We created a combined database consisting of four speech emotion-labeled databases, and we used it to train a 2D convolutional neural network to determine if the model could recognize multiple emotions in a speech sample. The model was able to classify the samples with an accuracy of 57.64\%. This shows that it is possible to apply machine learning to the problem of multi-label speech emotion recognition and to achieve a reasonable accuracy.\footnote{Code for this paper is available on Github at: \url{https://github.com/Brian-Pho/RVST598_Speech-Emotion-Recognition}}
\end{abstract}

%
% The code below should be generated by the tool at
% http://dl.acm.org/ccs.cfm
% Please copy and paste the code instead of the example below. 
%
\begin{CCSXML}
	<ccs2012>
	<concept>
	<concept_id>10010405.10010469.10010475</concept_id>
	<concept_desc>Applied computing~Sound and music computing</concept_desc>
	<concept_significance>500</concept_significance>
	</concept>
	</ccs2012>
\end{CCSXML}

\ccsdesc[500]{Applied computing~Sound and music computing}

\keywords{multi-label, speech, emotion, recognition, neural networks}

\maketitle

\input{paper_body}

\bibliographystyle{ACM-Reference-Format}
\bibliography{references}

\end{document}

### 2019-09-09 research/paper/paper_header.tex
Update paper with revisions and improvements

@@
 \begin{abstract}
-Current speech emotion recognition systems often overlook the multi-label data that comes with databases.
-In this paper, we address the problem of whether machine learning can be used to detect multiple emotions in speech.
-We created a combined database consisting of four speech emotion-labeled databases, and we used it to train a 2D convolutional neural network to determine if the model could recognize multiple emotions in a speech sample.
-The model was able to classify the samples with an accuracy of 57.64\%.
-This shows that it is possible to apply machine learning to the problem of multi-label speech emotion recognition and to achieve a reasonable accuracy.\footnote{Code for this paper is available on Github at: \url{https://github.com/Brian-Pho/RVST598_Speech-Emotion-Recognition}}
+A problem with current speech emotion recognition systems is that they can only recognize one emotion from a set of emotions in speech.
+This is a problem because it does not match how people recognize emotions in speech.
+When people hear speech, they can perceive multiple emotions instead of a single emotion.
+We address this problem by building a 2D convolutional neural network that can also recognize multiple emotions and it achieves an accuracy of 57.64\%.
+This model demonstrates a more realistic approach to speech emotion recognition by more closely matching human data.\footnote{Code for this paper is available on Github at: \url{https://github.com/Brian-Pho/RVST598_Speech-Emotion-Recognition}}
 \end{abstract}

### 2019-09-09 src/database_processor/display_cm.py

@@ -72,7 +72,7 @@ def visualize_confusion_matrix():
 
         fig.tight_layout()
         fig.set_size_inches(8, 8)
-        fig.savefig('{}.png'.format(em_label), dpi=600)
+        fig.savefig('{}.png'.format(em_label), dpi=300)
         # plt.show()
 
 

### 2019-09-22 research/paper AGM/paper_body.tex
Add FG paper format

--- 2019-09-07 research/paper/paper_body.tex
@@
 The field of affective computing studies the development of systems that can recognize, interpret, process, and simulate human affects.
-A subfield of affective computing that we are interested in is speech emotion recognition (SER).
-SER is the problem of recognizing which emotions are present in speech and SER has been approached in two ways:
+A subfield of affective computing, the one that we are interested in, is speech emotion recognition (SER).
+SER is the problem of recognizing which emotions are present in speech and it is important because of its applications to
+\begin{itemize}
+	\item call center conversations that improve service quality,
+	\item speech translation systems that are more accurate at conveying emotions,
+	\item and robotic pets that are more compassionate \cite{Koolagudi2012}.
+\end{itemize}
+
+\subsection{Literature Review}
+
+SER has historically been approached in two ways:
 \begin{itemize}
@@
 
-\subsection{Literature Review}
-
 Feature engineering is a method of manually extracting desired parameters from the data for use in  recognition.
-For audio, examples of parameters include pitch, energy, and Mel-Frequency Cepstral Coefficients (MFCC).\cite{Rybka2013} Nassif et al.
+For audio, examples of parameters include pitch, energy, and Mel-Frequency Cepstral Coefficients (MFCC) \cite{Rybka2013}.
+Nassif et al.
 \cite{Nassif2019} found that most researchers used the MFCC parameter for deep learning models but also recommended other parameters such as Linear Predictive Coding.
 
-Machine learning is a method of creating an artificial neural network that automatically extracts its own parameters for recognition.
-Examples include support vector machines (SVM), convolutional neural networks (CNN), and recurrent neural networks (RNN).
-Both approaches are not exclusive and a hybrid approach of feature engineering and machine learning results in better classification accuracy.\cite{Nassif2019}
+Machine learning is a method of creating models that automatically extract their own parameters for recognition without being manually programmed.
+Examples include convolutional neural networks (CNN), recurrent neural networks (RNN), and long short-term memory (LSTM).
+Both approaches are not exclusive and a hybrid approach of feature engineering and machine learning results in better classification accuracy \cite{Nassif2019}.
 
@@
 
-Current state-of-the-art SER has achieved an accuracy of 52.14\% for the case of speaker-independent, single-labeled emotions on the IEMOCAP database.\cite{Zhao2019} This result was achieved using a 2D CNN long short-term memory (LSTM) model where the audio waveform was converted into a log-Mel spectrogram and then fed into the model.
+Current state-of-the-art SER has achieved an accuracy of 52.14\% for the case of speaker-independent, single-labeled emotions on the IEMOCAP database \cite{Zhao2019}.
+This result was achieved using a 2D CNN LSTM model where the audio waveform was converted into a log-Mel spectrogram and then fed into the model.
 
@@
 The IEMOCAP \cite{busso_2008} and CREMA-D \cite{cao_2014} databases both include multiple labels for each speech sample but the data is discarded by considering the emotion with the majority of votes.
-We argue that discarding the non-majority votes results in a less realistic model of emotion classification due to not matching human performance and due to the loss of information by discarding votes.
+We argue that discarding the non-majority votes results in a less realistic model of emotion classification due to not matching human performance and due to the loss of information.
+For example, in the case of speech translation systems, translating "surprise" speech lacks the nuance of whether the surprise is "surprise-happy" such as the case of receiving a gift or if it is "surprise-sad" such as the case of receiving news of the death of a loved one.
+A single label is not able to capture this nuance compared to multiple labels and could result in inaccurate or offensive translations.
 
@@
 
-We approach the problem of multi-label emotion classification by using a 2D CNN to classify speech samples into multiple emotions.
+The main idea of this paper is to build a more realistic speech emotion recognition system that can recognize multiple emotions in speech.
+We approach this problem by using more data and by using a 2D CNN to classify speech samples into multiple emotions.
 We collected four databases: two with multi-label and single-label samples, and two with only single-label samples.
@@
 \begin{itemize}
-	\item Section II details the methodology we use such as the how we combined the four databases and describes the architecture of the neural network.
-	\item Section III describes and discusses the results from testing the neural network.
-	\item Section IV concludes this paper and provides suggestions for future directions.
+	\item Section 2 details the methodology we use such as the how we combined the four databases and describes the architecture of the neural network.
+	\item Section 3 describes and discusses the results from testing the neural network.
+	\item Section 4 concludes this paper and provides suggestions for future directions.
 \end{itemize}
@@
 \cite{Balakrishnan2017} found that CNN-based models have superior performance compared to RNNs.
-Another justification for using CNNs is due treating the log-Mel spectrograms as images which CNNs have been shown to perform well on.\cite{Krizhevsky2012}
+Another justification for using CNNs is due to treating the log-Mel spectrograms as images and CNNs have been shown to perform well on images \cite{Krizhevsky2012}.
 
-A high level overview of the data flow is shown in figure \ref{highLevelDataFlowDiagram} where samples flow through four stages of processing.
+A high level overview of the data flow is shown in figure \ref{highLevelDataFlowDiagram} where samples flow through three stages of processing.
 The stages are detailed further into the methodology.
-Both the preprocessing steps and the CNN were implemented in Python using the Librosa \cite{McFee2015} and Keras \cite{Chollet2015} library respectively.
+Both the preprocessing steps and the CNN were implemented in Python using the Librosa \cite{McFee2015} and Keras \cite{Chollet2015} libraries respectively.
 
-\subsection{Preprocessing}
+\subsection{Preprocessing} \label{Preprocessing}
 
@@
 
-To determine if a ML model can perform multi-label speech emotion recognition, we first need to obtain speech samples with their labeled emotion.
-We considered eight databases and chose four based on accessibility and the number of overlapping emotions.
+To build a more realistic SER system, we first need to obtain speech samples with their labeled emotion.
+We considered eight databases and chose four based on accessibility and based on the number of overlapping emotions.
 Table \ref{dbEmoTable} compares the set of emotions of each database which makes it easier to identify overlapping emotions.
@@
 We chose to mix of single- and multi- labeled data to increase the number of samples that the model can learn from and because the single-labels can be considered as special cases of multi-labeled data.
-However, we removed samples that were labeled with four or five emotions because we consider samples with that many emotions to be ambiguous and because outliers hinder a neural networks ability to learn.
-Outliers can cause large gradient updates which prevents a model from converging.
+However, we removed samples that were labeled with four or five emotions because we consider these samples to be ambiguous and because outliers can hinder a neural network's ability to learn.
+Outliers can hinder a neural network's ability to learn by causing large gradient updates that prevent the model from converging \cite{Chollet2017}.
 
@@
 
-Each sample from a database would flow through the same preprocessing steps to maintain consistency across all samples. The detailed steps are described below.
+Each sample from a database went through the same preprocessing steps to maintain consistency across all samples. The detailed steps are described below.
 \begin{enumerate}
-	\item A sample starts as a raw waveform in the form of time series points specifying the amplitude at a certain point in time.
-	\item The sample is then padded or cropped to the desired length of 4.5s. Shorter samples were zero-padded on the right tail. Longer samples were cropped and the extra information discarded.
+	\item A sample starts as a raw waveform in the form of time series points specifying the amplitude at a point in time.
+	\item The sample is then padded or cropped to the desired length of 4.5s.
+Shorter samples were zero-padded on the right tail.
+Longer samples were right-tail cropped and the extra information was discarded.
 	\item If the sample came from a database that we considered noisy, then a noise reduction filter was applied to the sample. We consider the IEMOCAP and CREMA-D databases to be noisy.
-	\item The sample is then converted into a log-Mel spectrogram using the short-time Fourier transform (STFT) and Mel scale equation.
-The phase information was discard as it does not seem to hold relevant information.
-\cite{Kozakowski2017}
+	\item The sample is then converted into a log-Mel spectrogram using the STFT and Mel scale.
+The phase information was discarded as it does not appear to hold relevant information \cite{Kozakowski2017}.
 	\item The final step is to normalize the spectrograms to have values between negative one and one. This was done by using a min-max scaling function.
 \end{enumerate}
-
-The discrete-time STFT is defined in (\ref{stft}) where $x[n]$ is the discrete signal and $w[n]$ is the window function.
 
@@
 
-The Mel scale is defined in (\ref{melScale}) where $f$ is the frequency in hertz and $m$ is mels.\cite{Oshaughnessy1990}
+The discrete-time STFT is defined in (\ref{stft}) where $x[n]$ is the discrete signal and $w[n]$ is the window function.
+We used a window size of 3072 with a 75\% overlap.
+This choice is based on the work of Zhao et al.
+\cite{Zhao2019} where they also use 75\% overlap but with a smaller window size (2048) and they achieved excellent results.
 
@@
 
-For the STFT, we used a window size of 3072 with a 75\% overlap.
-This choice is based on the work of Zhao et al.
-\cite{Zhao2019} where they also use 75\% overlap but with a smaller window size (2048) and they achieved excellent results.
-For the Mel spectrogram, we set the minimum frequency to 20 Hz and maximum frequency to 12 kHz with 200 Mel bins.
+The Mel scale is defined in (\ref{melScale}) where $f$ is the frequency in hertz and $m$ is mels \cite{Oshaughnessy1990}.
+We set the minimum frequency to 20 Hz and maximum frequency to 12 kHz with 200 Mel bins.
 The frequency range was chosen after experimenting with various ranges and selecting the range that resulted in visually clean spectrograms.
 The number of Mel bins was also chosen after experimentation as too few bins resulted in poor temporal resolution while too many bins resulted in poor frequency resolution.
-This is due to the Gabor limit which describes the trade-off between temporal and frequency resolution.
+
+\begin{equation}
+\label{minmaxscale}
+\begin{aligned}
+X_{std} &= \frac{X - X_{min}}{X_{max} - X_{min}}\\
+X_{scaled} &= X_{std} \cdot  (max - min) + min
+\end{aligned}
+\end{equation}
+
+The min-max scaling is defined in (\ref{minmaxscale}) where $X_{min}$ is the smallest value in tensor $X$, $X_{max}$ is the largest value in tensor $X$, $min$ is the lower bound set to $-1$, and $max$ is the upper bound set to $1$.
+We scaled all spectrograms to only have values between $-1$ and $1$ as this reduces the chances of triggering large gradient updates that prevent the model from converging \cite{Chollet2017}.
 
@@
 	\includegraphics[width=\textwidth]{neural_network_architecture.png}
-	\caption{Architecture for the speech emotion recognition network including the input layer and eight processing layers.}
+	\caption{Architecture for the SER network including the input layer and eight processing layers.}
 	\label{neuralNetworkArchitecture}
@@
 
-After all of the databases were processed this way, the final combined database was fed into a neural network for training, validation, and testing.
-We constructed an eight layer neural network consisting of four convolutional layers and three dense layers.
+After all of the databases were processed as described in section \ref{Preprocessing}, the final combined database was fed into a neural network for training, validation, and testing.
+We constructed an eight layer neural network consisting of four convolutional layers and four dense layers.
 The network architecture is shown in figure \ref{neuralNetworkArchitecture}.
@@
 	\item We split the combined database into 80\% training, 10\% validation, and 10\% testing.
-	\item We applied dropout to the dense layers and batch normalization to the convolutional layers to deal with overfitting. \cite{Srivastava2014}, \cite{Ioffe2015}
+	\item We applied dropout to the dense layers and batch normalization to the convolutional layers to deal with overfitting \cite{Srivastava2014}, \cite{Ioffe2015}.
 	\item We updated the model's hyperparameters based on the validation loss and accuracy to improve the model's accuracy and ability to generalize.
@@
 
-Table \ref{modelHyperparams} describes the network hyperparameters and table \ref{trainHyperparams} describes the training hyperparameters.
+Table \ref{modelHyperparams} describes the model's hyperparameters and table \ref{trainHyperparams} describes the training hyperparameters.
+The decision to use \textit{binary cross-entropy} as the loss measure is due to multi-label problems considering each output as independent and thus each output acts like a binary classifier.
+This lets the model only penalize the outputs that are incorrect without penalizing the other outputs.
+However, we use \textit{categorical cross-entropy} as the accuracy metric because binary cross-entropy would provide the incorrect accuracy as it defined for a binary output such as 0 or 1, not vectors containing binary outputs.
+We could have picked a different accuracy metric but chose categorical cross-entropy due to its ease of use.
+
+Likewise, we use \textit{sigmoid} as the output layer's activation function instead of the more commonly used \textit{softmax} function in classification.
+The justification is the same as in our choice of loss measure; sigmoid treats each output as independent which means multiple outputs can be true while softmax maximizes a single output while minimize all other outputs.
 
@@
 		Optimization Algorithm & Adam \\
+		Learning Rate & 0.001 \\
+		Beta 1 & 0.9 \\
+		Beta 2 & 0.999 \\
+		Epsilon & $10^{-8}$ \\
 		Loss Measure & Binary Cross-entropy \\
 		Accuracy Metric & Categorical Cross-entropy \\
-		Activation Function & Sigmoid \\
+		Hidden Layer Activation Function & ReLU \\
+		Output Layer Activation Function & Sigmoid \\
 		\bottomrule
@@
 
-The final model was evaluated on the testing set by using average accuracy.
-The average accuracy is calculated from the confusion matrices shown in figure \ref{confusionMatrix}.
-The confusion matrices are calculated by comparing the true labels to the predicted labels for each emotion.
-There is a confusion matrix for each emotion due to the multi-label problem considering each emotion as independent.
+The final model was evaluated on the testing set by using average accuracy that was calculated from the confusion matrices shown in figure \ref{confusionMatrix}.
+The confusion matrices are calculated by comparing the true labels to the predicted labels for each independent emotion.
 
@@
 Confusion matrices for each emotion are shown in Figure \ref{confusionMatrix} and comparisons to the current literature is shown in table \ref{litacccompare}.
+We compare our results to the single-label case of SER as we could not find papers with results for the multi-label case.
 
@@
 
-\begin{table}
+\begin{table*}
 	\centering
@@
 	\label{litacccompare}
-	\begin{tabular}{@{}lrr@{}}
+	\begin{tabular}{@{}llllllr@{}}
 		\toprule
-		Research Work & Method & Accuracy (\%) \\
+		Research Work & Method & Testing Method & Number of Emotions & Databases Used & Label Type & Accuracy (\%) \\
 		\midrule
-		Zhao et al. \cite{Zhao2019} & CNN + LSTM & 52.1 \\
-		Our Work & CNN & 57.6 \\
-		Etienne et al. \cite{Etienne2018} & CNN + LSTM & 61.7 \\
-		Zhang et al. \cite{Zhang2019} & CNN & 63.9 \\
-		Fayek et al. \cite{Fayek2017} & CNN & 64.8 \\
-		Yenigalla et al. \cite{Yenigalla2018} & CNN & 73.9 \\
-		Badshah et al. \cite {Badshah2019} & CNN & 84.3 \\
+		Zhao et al. \cite{Zhao2019} & CNN + LSTM & Testing set & 6 & IEMOCAP & Single & 52.1 \\
+		Our Work & CNN & Testing set & 7 & \begin{tabular}[c]{@{}l@{}}IEMOCAP, TESS, \\ RAVDESS, CREMA-D\end{tabular} & Multi & 57.6 \\
+		Etienne et al. \cite{Etienne2018} & CNN + LSTM & Testing set & 4 & IEMOCAP & Single & 61.7 \\
+		Zhang et al. \cite{Zhang2019} & CNN & Testing set & 4 & IEMOCAP & Single & 63.9 \\
+		Fayek et al. \cite{Fayek2017} & CNN & Testing set & 4 & IEMOCAP & Single & 64.8 \\
+		Yenigalla et al. \cite{Yenigalla2018} & CNN & Testing set & 4 & IEMOCAP & Single & 73.9 \\
+		Badshah et al. \cite {Badshah2019} & CNN & Testing set & 7 & Emo-DB, Korean dataset & Single & 80.8 \\
 		\bottomrule
 	\end{tabular}
-\end{table}
+\end{table*}
 
-The results achieved in this paper do not reach the state-of-the-art accuracy achieved for the single-label case but this is the first attempt on the harder case of mutli-label using more data.
-So we conclude that it is possible to achieve a reasonable accuracy on the problem of multi-label speech emotion recognition by applying deep learning to the problem.
+The results achieved in this paper do not reach the state-of-the-art accuracy achieved for the single-label case.
+But considering how this is the first attempt on the more realistic problem of multi-label SER, we perform reasonably well.
+Compared to the literature shown in table \ref{litacccompare}, no other paper considers as many emotions as we do while also using four databases and also tackling the multi-label problem.
+So we conclude that we are successful in building a more realistic SER system by using ML and more data to tackle the problem of recognizing multiple emotions in speech.
 
@@
 
-One limitation of this work is that we only accounted for seven emotions but recent research has suggested that there are more emotions such as boredom, shame, and triumph.
-\cite{Cordaro2018} However, one issue with expanding the set of considered emotions is the lack of databases with the labeled emotion.
+One limitation of this work is that we only accounted for seven emotions but recent research has suggested that there are more emotions such as boredom, shame, and triumph \cite{Cordaro2018}.
+However, one issue with expanding the set of considered emotions is the lack of databases with the labeled emotion.
 
 Another limitation of this work is that all samples are spoken in English so the model is biased towards Anglophones.
-In theory, the basic emotions are universal across languages and cultures so incorporating databases spoken in different languages, such as the Emo-DB database, would help the model generalize across languages.\cite{Burkhardt2005}
+In theory, the basic emotions are universal across languages and cultures so incorporating databases spoken in different languages, such as the Emo-DB database, would help the model generalize across languages \cite{Burkhardt2005}.
 

### 2019-09-22 research/paper FG/paper.tex
Add FG paper format

--- 2019-09-22 research/paper AGM/paper_body.tex
@@
-\section{Introduction}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%2345678901234567890123456789012345678901234567890123456789012345678901234567890
+%        1         2         3         4         5         6         7         8
+%
+% Slightly modified by X. Baró for FG2020
+%
+
+%\documentclass[letterpaper, 10 pt, conference]{ieeeconf}  % Comment this line out
+                                                          % if you need a4paper
+\documentclass[a4paper, 10pt, conference]{ieeeconf}      % Use this line for a4
+                                                          % paper
+\usepackage{FG2020}
+
+%\FGfinalcopy % *** Uncomment this line for the final submission
+
+\IEEEoverridecommandlockouts                              % This command is only
+                                                          % needed if you want to
+                                                          % use the \thanks command
+\overrideIEEEmargins
+% See the \addtolength command later in the file to balance the column lengths
+% on the last page of the document
+
+% The following packages can be found on http:\\www.ctan.org
+%\usepackage{graphics} % for pdf, bitmapped graphics files
+\usepackage{booktabs}
+\usepackage{graphicx}
+\graphicspath{{figures/}}
+\usepackage{hyperref}
+\usepackage{cite}
+
+%\usepackage{epsfig} % for postscript graphics files
+%\usepackage{mathptmx} % assumes new font selection scheme installed
+%\usepackage{times} % assumes new font selection scheme installed
+\usepackage{amsmath} % assumes amsmath package installed
+%\usepackage{amssymb}  % assumes amsmath package installed
+
+\def\FGPaperID{****} % *** Enter the FG2020 Paper ID here
+
+\title{\LARGE \bf
+Multi-Label Speech Emotion Recognition Using 2D Convolutional Neural Networks
+}
+
+%use this in case of a single affiliation
+\author{\parbox{16cm}{\centering
+    {\large Brian Pho, Thomas Truong, and Svetlana Yanushkevich}\\
+    {\normalsize
+    Biometric Technologies Laboratory, Department of Electrical and Computer Engineering, University of Calgary, Alberta, Canada\\}}
+}
+
+%use this in case of several affiliations
+%\author{\parbox{16cm}{\centering
+%    {\large Huibert Kwakernaak$^1$ and Pradeep Misra$^2$}\\
+%    {\normalsize
+%    $^1$ Faculty of Electrical Engineering, Mathematics and Computer Science, University of Twente, Enschede, The Netherlands\\
+%    $^2$ Department of Electrical Engineering, Wright State University, Dayton, USA}}
+%    \thanks{This work was not supported by any organization}% <-this % stops a space
+%}
+
+\begin{document}
+
+\ifFGfinal
+\thispagestyle{empty}
+\pagestyle{empty}
+\else
+\author{Anonymous FG2020 submission\\ Paper ID \FGPaperID \\}
+\pagestyle{plain}
+\fi
+\maketitle
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\begin{abstract}
+
+Current speech emotion recognition systems often overlook the multi-label data that comes with databases.
+In this paper, we address the problem of whether machine learning can be used to detect multiple emotions in speech.
+We created a combined database consisting of four speech emotion-labeled databases, and we used it to train a 2D convolutional neural network to determine if the model could recognize multiple emotions in a speech sample.
+The model was able to classify the samples with an accuracy of 57.64\%.
+This shows that it is possible to apply machine learning to the problem of multi-label speech emotion recognition and to achieve a reasonable accuracy.\footnote{Code for this paper is available on Github at: \url{https://github.com/Brian-Pho/RVST598_Speech-Emotion-Recognition}}
+
+\end{abstract}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{INTRODUCTION}
 
@@
 A subfield of affective computing, the one that we are interested in, is speech emotion recognition (SER).
-SER is the problem of recognizing which emotions are present in speech and it is important because of its applications to
-\begin{itemize}
-	\item call center conversations that improve service quality,
-	\item speech translation systems that are more accurate at conveying emotions,
-	\item and robotic pets that are more compassionate \cite{Koolagudi2012}.
-\end{itemize}
+SER is the problem of recognizing which emotions are present in speech and it is important because of its applications to: call center conversations that improve service quality, speech translation systems that are more accurate at conveying emotions, and robotic pets that are more compassionate \cite{Koolagudi2012}.
 
@@
 
-SER has historically been approached in two ways:
-\begin{itemize}
-	\item Feature engineering
-	\item Machine learning (ML)
-\end{itemize}
+SER has historically been approached in two ways: feature engineering and machine learning (ML).
 
@@
 
-\section{Methodology}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{METHODOLOGY}
 
@@
 
-\begin{table*}
-	\centering
-	\caption{Comparison of databases with their labeled emotions.}
-	\label{dbEmoTable}
-	\begin{tabular}{@{}lccccccccccccc@{}}
-		\toprule
-		Databases & \multicolumn{13}{c}{Emotions} \\ 
-		\cmidrule{2-14}
-		& Neutral & Anger & Disgust & Fear & Happy & Sad & Surprise & Calm & Excitement & Frustration & Amused & Sleepy & Bored \\ 
-		\midrule
-		IEMOCAP & x & x & x & x & x & x & x &   & x & x &   &   &   \\
-		CREMA-D & x & x & x & x & x & x &   &   &   &   &   &   &   \\
-		TESS & x & x & x & x & x & x & x &   &   &   &   &   &   \\
-		RAVDESS & x & x & x & x & x & x & x & x &   &   &   &   &   \\
-		MSP-IMPROV & x & x & x & x & x & x & x &   &   &   &   &   &   \\
-		SAVEE & x & x & x & x & x & x & x &   &   &   &   &   &   \\
-		Emo-DB & x & x & x & x & x & x &   &   &   &   &   &   & x \\
-		EmoV-DB & x & x & x &   &   &   &   &   &   &   & x & x &   \\
-		\bottomrule
-	\end{tabular}%
-\end{table*}
+%\begin{table*}
+%	\centering
+%	\caption{Comparison of databases with their labeled emotions.}
+%	\label{dbEmoTable}
+%	\begin{tabular}{@{}lccccccccccccc@{}}
+%		\toprule
+%		Databases & \multicolumn{13}{c}{Emotions} \\ 
+%		\cmidrule{2-14}
+%		& Neutral & Anger & Disgust & Fear & Happy & Sad & Surprise & Calm & Excitement & Frustration & Amused & Sleepy & Bored \\ 
+%		\midrule
+%		IEMOCAP & x & x & x & x & x & x & x &   & x & x &   &   &   \\
+%		CREMA-D & x & x & x & x & x & x &   &   &   &   &   &   &   \\
+%		TESS & x & x & x & x & x & x & x &   &   &   &   &   &   \\
+%		RAVDESS & x & x & x & x & x & x & x & x &   &   &   &   &   \\
+%		MSP-IMPROV & x & x & x & x & x & x & x &   &   &   &   &   &   \\
+%		SAVEE & x & x & x & x & x & x & x &   &   &   &   &   &   \\
+%		Emo-DB & x & x & x & x & x & x &   &   &   &   &   &   & x \\
+%		EmoV-DB & x & x & x &   &   &   &   &   &   &   & x & x &   \\
+%		\bottomrule
+%	\end{tabular}%
+%\end{table*}
 
-To build a more realistic SER system, we first need to obtain speech samples with their labeled emotion.
+To build a more realistic SER system, we first collect speech samples with their labeled emotion.
 We considered eight databases and chose four based on accessibility and based on the number of overlapping emotions.
-Table \ref{dbEmoTable} compares the set of emotions of each database which makes it easier to identify overlapping emotions.
 We chose four databases because it becomes more difficult to maintain consistency due to database variability.
@@
 \end{itemize}
-Given this variability, we chose the following four databases
-\begin{enumerate}
-	\item IEMOCAP \cite{busso_2008}
-	\item TESS \cite{dupuis_2011}
-	\item RAVDESS \cite{livingstone_2018}
-	\item CREMA-D \cite{cao_2014}
-\end{enumerate}
+Given this variability, we chose the following four databases: IEMOCAP \cite{busso_2008}, TESS \cite{dupuis_2011}, RAVDESS \cite{livingstone_2018}, and CREMA-D \cite{cao_2014}.
 
-To control for the set and number of labeled emotions, we consider the following seven emotions for recognition
-\begin{enumerate}
-	\item Neutral
-	\item Anger
-	\item Disgust
-	\item Fear
-	\item Happy
-	\item Sad
-	\item Surprise
-\end{enumerate}
+To control for the set and number of labeled emotions, we consider the following seven emotions for recognition: neutral, anger, disgust, fear, happy, sad, and surprise.
 We chose these seven emotions due to them being considered basic emotions by Ekman \cite{Ekman1992} and due to these seven being the most common among all databases.
@@
 
-We combined these four databases into a combined database to train, validate, and test the neural network. The combined database is detailed in figure \ref{combinedDb}.
-
-\begin{figure*}
-	\centering
-	\hspace{6mm}
-	\includegraphics[width=\textwidth]{combined_db.png} 
-	\caption{Proportions of each database, emotion, and label types in the combined database.}
-	\label{combinedDb}
-\end{figure*}
-
-Each sample from a database went through the same preprocessing steps to maintain consistency across all samples. The detailed steps are described below.
+We combined these four databases into a combined database to train, validate, and test the neural network.
+Each sample from a database went through the same preprocessing steps to maintain consistency across all samples.
+The detailed steps are described below.
 \begin{enumerate}
@@
 
-\begin{equation}
-\label{stft}
-\mathbf{STFT}\{x[n]\}(m,\omega) = \sum_{n=-\infty}^{\infty}x[n]w[n-m]e^{-j \omega n}
-\end{equation}
-
-The discrete-time STFT is defined in (\ref{stft}) where $x[n]$ is the discrete signal and $w[n]$ is the window function.
-We used a window size of 3072 with a 75\% overlap.
+For the STFT, we used a window size of 3072 with a 75\% overlap.
 This choice is based on the work of Zhao et al.
@@
 
-\begin{equation}
-\label{melScale}
-m = 2595 log_{10}(1 + \frac{f}{700})
-\end{equation}
-
-The Mel scale is defined in (\ref{melScale}) where $f$ is the frequency in hertz and $m$ is mels \cite{Oshaughnessy1990}.
-We set the minimum frequency to 20 Hz and maximum frequency to 12 kHz with 200 Mel bins.
+For the Mel scale, we set the minimum frequency to 20 Hz and maximum frequency to 12 kHz with 200 Mel bins.
 The frequency range was chosen after experimenting with various ranges and selecting the range that resulted in visually clean spectrograms.
@@
 
-\begin{figure}
-	\centering
-	\hspace{6mm}
-	\includegraphics[width=\linewidth]{high_level_dataflow_diagram.png}
-	\caption{High level overview of the processing stages that a speech sample goes through.}
-	\label{highLevelDataFlowDiagram}
-\end{figure}
+%\begin{figure}
+%	\centering
+%	\hspace{6mm}
+%	\includegraphics[width=\linewidth]{high_level_dataflow_diagram.png}
+%	\caption{High level overview of the processing stages that a speech sample goes through.}
+%	\label{highLevelDataFlowDiagram}
+%\end{figure}
 
@@
 
-\begin{figure*}
-	\centering
-	\hspace{6mm}
-	\includegraphics[width=\textwidth]{neural_network_architecture.png}
-	\caption{Architecture for the SER network including the input layer and eight processing layers.}
-	\label{neuralNetworkArchitecture}
-\end{figure*}
+%\begin{figure*}
+%	\centering
+%	\hspace{6mm}
+%	\includegraphics[width=\textwidth]{neural_network_architecture.png}
+%	\caption{Architecture for the SER network including the input layer and eight processing layers.}
+%	\label{neuralNetworkArchitecture}
+%\end{figure*}
 
@@
 We constructed an eight layer neural network consisting of four convolutional layers and four dense layers.
-The network architecture is shown in figure \ref{neuralNetworkArchitecture}.
 The training process is described below.
@@
 
-\section{Results and Discussion}
+\addtolength{\textheight}{-3cm}   % This command serves to balance the column lengths
+                                  % on the last page of the document manually. It shortens
+                                  % the textheight of the last page by a suitable amount.
+                                  % This command does not take effect until the next page
+                                  % so it should come on the page before the last. Make
+                                  % sure that you do not shorten the textheight too much.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{RESULTS AND DISCUSSION}
 
@@
 The final accuracy achieved was 57.64\%.
-Confusion matrices for each emotion are shown in Figure \ref{confusionMatrix} and comparisons to the current literature is shown in table \ref{litacccompare}.
+Comparisons to the current literature is shown in table \ref{litacccompare}.
 We compare our results to the single-label case of SER as we could not find papers with results for the multi-label case.
 
-\begin{figure*}
+\begin{figure}
 	\centering
 	\hspace{6mm}
-	\includegraphics[width=\textwidth]{confusion_matrix.png}
+	\includegraphics[width=\linewidth]{confusion_matrix.png}
 	\caption{Confusion matrices for each emotion.}
 	\label{confusionMatrix}
-\end{figure*}
+\end{figure}
 
@@
 
-\section{Conclusion and Future Work}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{CONCLUSION AND FUTURE WORK}
 
@@
 
-\begin{acks}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{ACKNOWLEDGMENTS}
+
 This work was funded by the Program for Undergraduate Research Experience (PURE) award granted by the University of Calgary.
-\end{acks}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\bibliographystyle{IEEEtran}
+\bibliography{IEEEabrv,references}
+
+\end{document}

### 2019-09-24 research/paper FG/paper.tex
Updates to FG paper

@@
 \usepackage{cite}
-
 %\usepackage{epsfig} % for postscript graphics files
@@
 
-Current speech emotion recognition systems often overlook the multi-label data that comes with databases.
-In this paper, we address the problem of whether machine learning can be used to detect multiple emotions in speech.
-We created a combined database consisting of four speech emotion-labeled databases, and we used it to train a 2D convolutional neural network to determine if the model could recognize multiple emotions in a speech sample.
-The model was able to classify the samples with an accuracy of 57.64\%.
-This shows that it is possible to apply machine learning to the problem of multi-label speech emotion recognition and to achieve a reasonable accuracy.\footnote{Code for this paper is available on Github at: \url{https://github.com/Brian-Pho/RVST598_Speech-Emotion-Recognition}}
+A problem with current speech emotion recognition systems is that they can only recognize one emotion, from a set of emotions, in speech.
+This is a problem because it does not match how people recognize emotions in speech.
+When people hear speech, they can perceive multiple emotions instead of a single emotion.
+We address this problem by building a 2D convolutional neural network that can also recognize multiple emotions and it achieves an accuracy of 57.64\%.
+This model demonstrates a more realistic approach to speech emotion recognition by more closely matching human data.
+%\footnote{Code for this paper is available on Github at: \url{https://github.com/Brian-Pho/RVST598_Speech-Emotion-Recognition}}
 
@@
 
-SER has historically been approached in two ways: feature engineering and machine learning (ML).
-
-Feature engineering is a method of manually extracting desired parameters from the data for use in  recognition.
+SER has historically been approached in two ways: feature engineering (FE) and machine learning (ML).
+FE is a method of manually extracting desired parameters from the data for use in recognition.
 For audio, examples of parameters include pitch, energy, and Mel-Frequency Cepstral Coefficients (MFCC) \cite{Rybka2013}.
@@
 
-Machine learning is a method of creating models that automatically extract their own parameters for recognition without being manually programmed.
+ML is a method of creating models that automatically extract their own parameters for recognition without being manually programmed.
 Examples include convolutional neural networks (CNN), recurrent neural networks (RNN), and long short-term memory (LSTM).
-Both approaches are not exclusive and a hybrid approach of feature engineering and machine learning results in better classification accuracy \cite{Nassif2019}.
+Both approaches are not exclusive and a hybrid approach of FE and ML results in better classification accuracy \cite{Nassif2019}.
 
 A common example of combining both approaches is to convert an audio waveform into a spectrogram using the Short-Time Fourier Transform (STFT) and to feed the spectrogram into a neural network.
-Papers such as \cite{Engel2019}, \cite{Chen2018}, \cite{Badshah2019}, and \cite{Zhao2019} have demonstrated that this approach is successful for processing audio and for recognizing emotions in speech.
+Papers such as \cite{Engel2019}, \cite{Chen2018}, \cite{Badshah2019}, and \cite{Zhao2019} have demonstrated that this approach is successful for processing audio and for SER.
 
-Current state-of-the-art SER has achieved an accuracy of 52.14\% for the case of speaker-independent, single-labeled emotions on the IEMOCAP database \cite{Zhao2019}.
+Current state-of-the-art SER has achieved an accuracy of 52.14\% for the case of speaker-independent, single-label emotions on the IEMOCAP database \cite{Zhao2019}.
 This result was achieved using a 2D CNN LSTM model where the audio waveform was converted into a log-Mel spectrogram and then fed into the model.
 
-However, this model and most other models only consider a single emotion per speech sample and are often only trained on one or two databases.
+However, this model, and other current models, only consider a single emotion per speech sample and are often only trained on one or two databases.
 Kim et al.
-\cite{Kim2018a} is an exception as they consider the multi-label case of SER but they did not use machine learning and they used visual data on top of audio data.
-This paper extends upon current literature by considering the problem of multi-label SER using four databases to train and test a CNN.
+\cite{Kim2018a} is an exception as they consider the multi-label case of SER but they did not use ML and they used visual data on top of audio data.
+This paper extends upon the current literature by considering the problem of multi-label SER by using four databases to train and test a 2D CNN.
 
-The IEMOCAP \cite{busso_2008} and CREMA-D \cite{cao_2014} databases both include multiple labels for each speech sample but the data is discarded by considering the emotion with the majority of votes.
-We argue that discarding the non-majority votes results in a less realistic model of emotion classification due to not matching human performance and due to the loss of information.
-For example, in the case of speech translation systems, translating "surprise" speech lacks the nuance of whether the surprise is "surprise-happy" such as the case of receiving a gift or if it is "surprise-sad" such as the case of receiving news of the death of a loved one.
-A single label is not able to capture this nuance compared to multiple labels and could result in inaccurate or offensive translations.
+The IEMOCAP \cite{busso_2008} and CREMA-D \cite{cao_2014} databases both include multiple labels for each speech sample but the data is usually discarded by considering the emotion with the majority of votes.
+We argue that discarding the non-majority votes results in a less realistic SER model, due to not matching human performance and due to the loss of information.
+For example, in the case of speech translation systems, translating "surprise" speech lacks the nuance of whether the surprise is "surprise-happy" such as the case of receiving a gift, or if it is "surprise-sad" such as the case of receiving news of a tragedy.
+A single label is not able to capture this nuance compared to multiple labels and this could result in inaccurate or offensive translations.
 
@@
 
-The main idea of this paper is to build a more realistic speech emotion recognition system that can recognize multiple emotions in speech.
+The main idea of this paper is to build a more realistic SER system that can recognize multiple emotions in speech.
 We approach this problem by using more data and by using a 2D CNN to classify speech samples into multiple emotions.
-We collected four databases: two with multi-label and single-label samples, and two with only single-label samples.
-We then combined the four databases into an a single database by processing all of the speech samples into log-Mel spectrograms.
+We used four public databases then combined the databases into an a single database by processing all of the speech samples into log-Mel spectrograms.
 These log-Mel spectrograms were then fed into an eight-layer neural network consisting of four convolutional layers and four dense layers.
+The final output of the neural network is a list of present emotions.
 
@@
 \begin{itemize}
-	\item Section 2 details the methodology we use such as the how we combined the four databases and describes the architecture of the neural network.
+	\item Section 2 details the methodology we use such as how we combined the four databases and describes the architecture of the neural network.
 	\item Section 3 describes and discusses the results from testing the neural network.
@@
 We preprocess the speech samples into log-Mel spectrogram to help the neural network extract features relevant to emotion recognition.
-This choice is based on previous work such as \cite{Engel2019} and \cite{Badshah2017}.
-The choice of using a CNN is also based on previous work where Balakrishnan et al.
-\cite{Balakrishnan2017} found that CNN-based models have superior performance compared to RNNs.
+This choice is based on the previous work by \cite{Engel2019} and \cite{Badshah2017} where log-Mel spectrograms were successfully used for extracting audio features.
+
+The choice to use a CNN is based on the previous work of Balakrishnan et al.
+\cite{Balakrishnan2017} which found that CNN-based models have superior performance compared to RNNs.
 Another justification for using CNNs is due to treating the log-Mel spectrograms as images and CNNs have been shown to perform well on images \cite{Krizhevsky2012}.
-
-A high level overview of the data flow is shown in figure \ref{highLevelDataFlowDiagram} where samples flow through three stages of processing.
-The stages are detailed further into the methodology.
 Both the preprocessing steps and the CNN were implemented in Python using the Librosa \cite{McFee2015} and Keras \cite{Chollet2015} libraries respectively.
@@
 %\end{table*}
+%Table \ref{dbEmoTable} compares the set of emotions of each database which makes it easier to identify overlapping emotions.
 
@@
 We considered eight databases and chose four based on accessibility and based on the number of overlapping emotions.
-We chose four databases because it becomes more difficult to maintain consistency due to database variability.
+We chose four databases because it becomes more difficult to maintain consistency in the final samples due to database variability.
 Databases can vary in
@@
 
+\begin{figure}
+	\centering
+	\hspace{6mm}
+	\includegraphics[width=\linewidth]{neural_network_architecture.png}
+	\caption{Architecture for the SER network including the input layer and eight processing layers.}
+	\label{neuralNetworkArchitecture}
+\end{figure}
+
 After all of the databases were processed as described in section \ref{Preprocessing}, the final combined database was fed into a neural network for training, validation, and testing.
 We constructed an eight layer neural network consisting of four convolutional layers and four dense layers.
+The network architecture is shown in figure \ref{neuralNetworkArchitecture}.
 The training process is described below.
@@
 In analyzing the confusion matrices, we see that the model predicts an emotion is absent in most of the samples with "surprise" being the exception.
-We suspect that "surprise" being an exception is due to the class imbalance that is shown in figure \ref{combinedDb}.
+We suspect that "surprise" being an exception is due to the class imbalance.
 The "surprise" class was the least represented class out of the seven classes and this makes the model predict it more due to the use of class weights.
@@
 
-One limitation of this work is that we only accounted for seven emotions but recent research has suggested that there are more emotions such as boredom, shame, and triumph \cite{Cordaro2018}.
+One limitation of this work is that we only considered seven emotions but research has suggested that there are more emotions such as boredom, shame, and triumph \cite{Cordaro2018}.
 However, one issue with expanding the set of considered emotions is the lack of databases with the labeled emotion.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{ACKNOWLEDGMENTS}
-
-This work was funded by the Program for Undergraduate Research Experience (PURE) award granted by the University of Calgary.
+%\section{ACKNOWLEDGMENTS}
+%This work was funded by the PURE award granted by the University of Calgary.
 

### 2019-10-11 research/paper FG/paper.tex
Update paper
Update paper and submit to FG2020

@@
 }
-
-%use this in case of several affiliations
-%\author{\parbox{16cm}{\centering
-%    {\large Huibert Kwakernaak$^1$ and Pradeep Misra$^2$}\\
-%    {\normalsize
-%    $^1$ Faculty of Electrical Engineering, Mathematics and Computer Science, University of Twente, Enschede, The Netherlands\\
-%    $^2$ Department of Electrical Engineering, Wright State University, Dayton, USA}}
-%    \thanks{This work was not supported by any organization}% <-this % stops a space
-%}
 
@@
 We approach this problem by using more data and by using a 2D CNN to classify speech samples into multiple emotions.
-We used four public databases then combined the databases into an a single database by processing all of the speech samples into log-Mel spectrograms.
+We used four public databases and combined them into an a single database by processing all of the speech samples into log-Mel spectrograms.
 These log-Mel spectrograms were then fed into an eight-layer neural network consisting of four convolutional layers and four dense layers.
-The final output of the neural network is a list of present emotions.
+The final output of the neural network is a list of the emotions present in the speech sample.
 
@@
 We preprocess the speech samples into log-Mel spectrogram to help the neural network extract features relevant to emotion recognition.
-This choice is based on the previous work by \cite{Engel2019} and \cite{Badshah2017} where log-Mel spectrograms were successfully used for extracting audio features.
+This choice is based on the work by \cite{Engel2019} and \cite{Badshah2017} where log-Mel spectrograms were successfully used for extracting audio features.
 
-The choice to use a CNN is based on the previous work of Balakrishnan et al.
+The choice to use a CNN is based on the work of Balakrishnan et al.
 \cite{Balakrishnan2017} which found that CNN-based models have superior performance compared to RNNs.
@@
 \subsection{Preprocessing} \label{Preprocessing}
-
-%\begin{table*}
-%	\centering
-%	\caption{Comparison of databases with their labeled emotions.}
-%	\label{dbEmoTable}
-%	\begin{tabular}{@{}lccccccccccccc@{}}
-%		\toprule
-%		Databases & \multicolumn{13}{c}{Emotions} \\ 
-%		\cmidrule{2-14}
-%		& Neutral & Anger & Disgust & Fear & Happy & Sad & Surprise & Calm & Excitement & Frustration & Amused & Sleepy & Bored \\ 
-%		\midrule
-%		IEMOCAP & x & x & x & x & x & x & x &   & x & x &   &   &   \\
-%		CREMA-D & x & x & x & x & x & x &   &   &   &   &   &   &   \\
-%		TESS & x & x & x & x & x & x & x &   &   &   &   &   &   \\
-%		RAVDESS & x & x & x & x & x & x & x & x &   &   &   &   &   \\
-%		MSP-IMPROV & x & x & x & x & x & x & x &   &   &   &   &   &   \\
-%		SAVEE & x & x & x & x & x & x & x &   &   &   &   &   &   \\
-%		Emo-DB & x & x & x & x & x & x &   &   &   &   &   &   & x \\
-%		EmoV-DB & x & x & x &   &   &   &   &   &   &   & x & x &   \\
-%		\bottomrule
-%	\end{tabular}%
-%\end{table*}
-%Table \ref{dbEmoTable} compares the set of emotions of each database which makes it easier to identify overlapping emotions.
 
@@
 To control for the audio quality, we resampled all samples to 48 kHz, applied noise reduction to samples from the IEMOCAP and CREMA-D databases, and cropped all samples to 4.5 seconds.
-We used a kaiser filter for resampling and used spectral gating for noise reduction.
+We used a kaiser filter for resampling and we used spectral gating for noise reduction.
 We only applied noise reduction to the IEMOCAP and CREMA-D databases after listening to the audio.
@@
 
-%\begin{figure}
-%	\centering
-%	\hspace{6mm}
-%	\includegraphics[width=\linewidth]{high_level_dataflow_diagram.png}
-%	\caption{High level overview of the processing stages that a speech sample goes through.}
-%	\label{highLevelDataFlowDiagram}
-%\end{figure}
-
 \subsection{Neural Network}
-
-%\begin{figure*}
-%	\centering
-%	\hspace{6mm}
-%	\includegraphics[width=\textwidth]{neural_network_architecture.png}
-%	\caption{Architecture for the SER network including the input layer and eight processing layers.}
-%	\label{neuralNetworkArchitecture}
-%\end{figure*}
 
@@
 However, we use \textit{categorical cross-entropy} as the accuracy metric because binary cross-entropy would provide the incorrect accuracy as it defined for a binary output such as 0 or 1, not vectors containing binary outputs.
-We could have picked a different accuracy metric but chose categorical cross-entropy due to its ease of use.
+Different loss functions should be considered in the future to improve training results.
 
